---
title: "Deferred: campaign total and ordering"
---

<!-- Parked here until the report reaches these topics; underscore
prefix keeps this file out of the site render. Requires the setup
and per-game chunks from index.qmd (cum_scores builds on `scores`). -->

The campaign total is what decides the winner — the running sum of the
bank slips through 1898, then each post-campaign addition stepped in
one at a time: post cards (from the vaults), circus flyer, shares, and
finally the timetable. The number at the right is each company's final
score.

```{r cumulative}
#| fig-height: 6
cum_scores <- scores |>
  arrange(player, year) |>
  group_by(player) |>
  mutate(total = cumsum(dollars)) |>
  ungroup()

component_order <- c("postcards", "circus", "shares", "timetable")
component_x <- setNames(1901 + 3 * seq_along(component_order), component_order)

additions_wide <- dbGetQuery(con, "
  SELECT player, component, dollars FROM final_components
  WHERE component <> 'bankslips'
")

steps <- cum_scores |>
  filter(year == max(year)) |>
  select(player, total)
step_rows <- list()
for (comp in component_order) {
  steps <- steps |>
    left_join(filter(additions_wide, component == comp) |>
        select(player, add = dollars), by = "player") |>
    mutate(total = total + coalesce(add, 0)) |>
    select(-add)
  step_rows[[comp]] <- mutate(steps, year = component_x[[comp]])
}

cum_full <- bind_rows(
  select(cum_scores, player, year, total),
  bind_rows(step_rows)
) |>
  mutate(player = factor(player, levels = names(player_colors)))

totals_lab <- cum_full |>
  filter(year == max(year)) |>
  arrange(total) |>
  mutate(
    lab_y = total,
    # push overlapping labels apart (min 45 dollar-units of separation)
    lab_y = Reduce(
      function(prev, cur) max(cur, prev + 45),
      lab_y, accumulate = TRUE
    )
  )

ggplot(cum_full, aes(year, total, color = player)) +
  annotate("segment",
    x = 1865, xend = 1898, y = -95, yend = -95,
    color = "gray55", linewidth = 0.5
  ) +
  annotate("text",
    x = (1865 + 1898) / 2, y = -160, label = "bank slips",
    color = "gray45", size = 3.2
  ) +
  annotate("segment",
    x = min(component_x), xend = max(component_x), y = -95, yend = -95,
    color = "gray55", linewidth = 0.5
  ) +
  annotate("text",
    x = mean(range(component_x)), y = -160, label = "vault",
    color = "gray45", size = 3.2
  ) +
  geom_vline(xintercept = 1901, color = "gray80", linetype = "dotted") +
  annotate("text",
    x = 1900, y = 650, label = "post campaign",
    angle = 90, color = "gray55", size = 3
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_text(
    data = totals_lab,
    aes(y = lab_y, label = paste0("$", total)),
    hjust = -0.2, size = 3.3, fontface = "bold", show.legend = FALSE
  ) +
  scale_color_manual(values = player_colors) +
  scale_x_continuous(
    breaks = c(unique(scores$year), component_x),
    labels = c(unique(scores$year), "post cards", "circus flyer",
               "shares", "timetable"),
    expand = expansion(mult = c(0.02, 0.06))
  ) +
  coord_cartesian(clip = "off") +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    plot.margin = margin(10, 20, 26, 10)
  ) +
  labs(
    title = "Campaign total (bank slips, then each addition stepped in)",
    x = NULL, y = "cumulative dollars", color = NULL
  )
```

The post-campaign vault additions nearly rewrote the podium: green's $259 of
additions ($110 shares, $71 circus, $30 timetable, $48 post cards)
closed the gap to black from $66 to just **$18** — final score $1363 to
$1345.

```{r final-additions}
#| fig-height: 4
additions <- dbGetQuery(con, "
  SELECT player, component, dollars
  FROM final_components
  WHERE component <> 'bankslips'
") |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    component = factor(component,
      levels = c("shares", "circus", "timetable", "postcards")
    )
  )

ggplot(additions, aes(player, dollars, fill = component)) +
  geom_col(color = "white", linewidth = 0.3) +
  scale_fill_manual(values = c(
    shares = "#4477aa", circus = "#ee6677",
    timetable = "#228833", postcards = "#ccbb44"
  )) +
  labs(
    title = "Post-campaign vault additions by component",
    x = NULL, y = "dollars", fill = NULL
  )
```

### Did the ordering change?

Rank 1 = leading. First the rank within each individual game — extended
past 1898 with the rank within each vault addition (who earned the most
post cards, circus flyer, shares, timetable) and the final total. Then
the rank on the running campaign total as each of those steps in — the
race that matters.

```{r rank-helpers}
#| include: false
# sigmoid interpolation between rank points, sankey/bump style
bump_curves <- function(df, n = 40) {
  df <- arrange(df, player, year)
  do.call(rbind, lapply(split(df, df$player), function(d) {
    segs <- lapply(seq_len(nrow(d) - 1), function(i) {
      t <- seq(0, 1, length.out = n)
      s <- 1 / (1 + exp(-12 * (t - 0.5)))   # sigmoid easing
      data.frame(
        player = d$player[i],
        x = d$year[i] + t * (d$year[i + 1] - d$year[i]),
        y = d$rank[i] + s * (d$rank[i + 1] - d$rank[i])
      )
    })
    do.call(rbind, segs)
  }))
}

plot_bump <- function(rank_df, title, x_breaks, x_labels) {
  curves <- bump_curves(rank_df)
  ends <- filter(rank_df, year == max(year))
  ggplot() +
    geom_vline(xintercept = 1901, color = "gray80", linetype = "dotted") +
    geom_line(
      data = curves, aes(x, y, color = player, group = player),
      linewidth = 2.4, alpha = 0.85, lineend = "round"
    ) +
    geom_point(data = rank_df, aes(year, rank, color = player), size = 3.2) +
    geom_text(
      data = ends,
      aes(year, rank, label = player, color = player),
      hjust = -0.35, size = 3.4, fontface = "bold", show.legend = FALSE
    ) +
    scale_color_manual(values = player_colors, guide = "none") +
    scale_y_reverse(breaks = 1:5) +
    scale_x_continuous(
      breaks = x_breaks,
      labels = x_labels,
      expand = expansion(mult = c(0.02, 0.09))
    ) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1)) +
    labs(title = title, x = NULL, y = "rank")
}

# x positions for the vault steps + final total on the rank charts
rank_steps <- c("postcards", "circus", "shares", "timetable", "total")
rank_step_x <- setNames(1901 + 3 * seq_along(rank_steps), rank_steps)
rank_breaks <- c(unique(scores$year), rank_step_x)
rank_labels <- c(unique(scores$year), "post cards", "circus flyer",
                 "shares", "timetable", "total")
```

```{r ranks-game}
#| fig-height: 4
ranks <- cum_scores |>
  group_by(year) |>
  mutate(
    game_rank = rank(-dollars, ties.method = "min"),
    campaign_rank = rank(-total, ties.method = "min")
  ) |>
  ungroup()

# per-component ranks: who earned the most from each vault addition,
# then the final standings as 'total'
comp_ranks <- dbGetQuery(con, "
  SELECT player, component, dollars FROM final_components
  WHERE component <> 'bankslips'
") |>
  group_by(component) |>
  mutate(rank = rank(-dollars, ties.method = "min")) |>
  ungroup() |>
  transmute(player, year = rank_step_x[component], rank)

total_rank <- dbGetQuery(con, "SELECT player, place FROM v_final_standings") |>
  transmute(player, year = rank_step_x[["total"]], rank = place)

game_ranks_full <- bind_rows(
  transmute(ranks, player = as.character(player), year, rank = game_rank),
  comp_ranks,
  total_rank
) |>
  mutate(player = factor(player, levels = names(player_colors)))

plot_bump(
  game_ranks_full,
  "Rank within each game, each vault addition, and the final total",
  rank_breaks, rank_labels
)
```

```{r ranks-campaign}
#| fig-height: 4
# cumulative rank at every year and after each vault addition steps in
cum_ranks_full <- cum_full |>
  bind_rows(
    cum_full |>
      filter(year == max(year)) |>
      mutate(year = rank_step_x[["total"]])
  ) |>
  group_by(year) |>
  mutate(rank = rank(-total, ties.method = "min")) |>
  ungroup() |>
  transmute(player, year, rank)

plot_bump(
  cum_ranks_full,
  "Rank on running total (bank slips, then each vault addition)",
  rank_breaks, rank_labels
)
```

