---
title: "Ticket to Ride Legacy — Data Dive"
subtitle: "One campaign, 12 plays, 5 companies — reconstructed from the dead letter office"
---

```{r setup}
#| include: false
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)

con <- dbConnect(SQLite(), "ttr.db")

player_colors <- c(
  black  = "#1a1a1a",
  blue   = "#1f77c4",
  green  = "#2a9d3a",
  yellow = "#e6b400",
  red    = "#d62728"
)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold")
    )
)

scores <- dbGetQuery(con, "SELECT * FROM v_scores") |>
  mutate(player = factor(player, levels = names(player_colors)))

players <- dbGetQuery(con, "SELECT * FROM players")
```

## What on earth is *Ticket to Ride Legacy*?

[*Ticket to Ride*](https://boardgamegeek.com/boardgame/9209/ticket-to-ride)
(TtR) is Alan R. Moon's classic gateway board game: a set-collection
and network-building train game. Most players where I come from are
familiar with
[*Europe*](https://boardgamegeek.com/boardgame/14996/ticket-to-ride-europe)
or
[*Nordic Countries*](https://boardgamegeek.com/boardgame/31627/ticket-to-ride-nordic-countries),
but regardless of which *TtR* map you know, the core gameplay remains
largely the same: collect coloured train cards, claim routes between
cities, complete destination tickets for points, and score additional
bonuses from map-specific objectives. The player with the most points
at the end of the game wins.

[*Ticket to Ride Legacy: Legends of the West*](https://boardgamegeek.com/boardgame/390092/ticket-to-ride-legacy-legends-of-the-west)
(2023) gives that familiar formula the **legacy treatment** pioneered
by Rob Daviau and Matt Leacock in *Pandemic Legacy* (Seasons
[1](https://boardgamegeek.com/boardgame/161936/pandemic-legacy-season-1),
[2](https://boardgamegeek.com/boardgame/221107/pandemic-legacy-season-2)
and
[0](https://boardgamegeek.com/boardgame/314040/pandemic-legacy-season-0)
--- in that order). Across a twelve-game campaign, the outcome of each
session permanently changes future games. The map expands frontier by
frontier, new cards and mechanics are introduced, existing components
are modified with stickers or punches, and --- unlike many legacy
games --- almost nothing is destroyed. Instead, components eventually
retire to the **Dead Letter Office**, creating a persistent record of
the campaign as the story unfolds.

::: {.callout-warning title="Spoiler warning"}
This report is one long spoiler. It contains the complete campaign,
including every frontier, the fully revealed map, destination tickets,
postcards, hidden card values, campaign events, and final scoring. If
you think you might play *TtR Legacy: Legends of the West* in the
future, stop reading now and come back after twelve games.
:::

## Our campaign

Our campaign was played by five players. Throughout this report they
are referred to by their player colours --- **Green**, **Black**,
**Red**, **Yellow**, and **Blue**. Since each player controls a
railroad company throughout the campaign, the terms *player* and
*company* are used interchangeably.

The game itself was a Christmas present from Green and Black to
Black's sisters, Red and Yellow, and Yellow's husband, Blue. Everyone
at the table was already well versed in *TtR* ---
particularly *Europe* --- and are either budding hobbyists to
hard-core gamers. Only Green and Black had previous experience with
legacy games, having completed *Pandemic Legacy* Seasons 1 and 2 (with
Season 0 still on the shelf of opportunity), and they also have some
20+ plays of
[*Gloomhaven*](https://boardgamegeek.com/boardgame/174430/gloomhaven)
under their belt (still unfinished, but the dragon is nigh). As
*Pandemic Legacy* and *Gloomhaven* are cooperative games in which
players work together against the game, *Legends of the West* was
everyone's **first competitive legacy campaign**.

Although the campaign consists of only twelve games, ours spanned
roughly **two and a half years**. The game was received as a Christmas
present in 2023 and the first session took place in February 2024, but
after several games the campaign paused from late 2024 until April
2026. In the meantime, life happened: pregnancies, the arrival of two
new family members in 2025, and the usual challenge of finding time to
get everyone back around the table.

After the campaign concluded, the complete game state was
painstakingly **reconstructed by hand**. Every remaining card in the
Storeroom, Post Office, Company Vaults, and the entire Dead Letter
Office was catalogued card by card and entered into the database on
which the analyses in this report are based.

## Campaign basics

*Legends of the West* is played over twelve games spanning the years
1865 to 1898, with each session advancing the calendar by three years
(i.e., `1865:3:1898`). Every player controls a railroad company whose
progress persists throughout the campaign. Between games, the map
expands westward through newly opened **Frontiers** (jigsaw pieces
that gradually form the board), while new cards, mechanics, and
scoring opportunities themed to each frontier are introduced. As a
result, every campaign unfolds differently, as the order in which
these mechanics appear depends on the sequence in which the frontiers
are explored.

Unlike a traditional game of *TtR*, companies develop asymmetrically
over time. **Employees** provide unique abilities for a single game
(my personal favourite is the *Financier*). The game is seeded with
**Big Cities** (with more added during the campaign); every time a
company builds a route into a Big City, it draws one bonus train card
from the face-down deck (but only one bonus card per turn --- even if
a single route connects two Big Cities at once). Company **Headquarters** ---
permanent stickers placed on small city markers --- create ongoing
sources of income for their owners. The campaign's recurring villain,
**Mama O'Connell**, has a habit of collecting money that would
otherwise be yours: she can establish headquarters of her own, and she
frequently appears in the event deck demanding payment, while other
events may reward companies that meet certain conditions. Components,
primarily cards, are never destroyed. Instead, they may be modified
and stored for future games until they are eventually retired to the
**Dead Letter Office** --- or, in some cases, simply depleted, such as
a fully used sticker sheet.

Unlike traditional *TtR*, *Legends of the West* tracks **dollars**
rather than victory points. Money is earned from completed destination
tickets, coins remaining in hand, bonuses for trains left at the end
of the game, favourable events, and the $2 toll other companies pay
when building into one of your headquarters cities. Money is lost
through failed destination tickets, unfavourable events, payments
claimed by Mama O'Connell, and payments claimed by other companies
(note you can also pay these tolls with train cards instead). Every
game begins with each company holding $5 (although employee abilities may increase this starting amount).

::: {.callout-note title="Why 1865–1898?" collapse="true"}
The campaign follows the great American railroad boom. It begins in
1865, the year the American Civil War ended, and concludes in 1898 as
the nineteenth century --- and the era of westward railroad
expansion --- draws to a close. The rulebook deliberately leans into
familiar themes of frontier expansion and the Wild West, while also
acknowledging that this history came at a significant human cost
through the dispossession of Native American peoples and the
exploitation of railroad labourers. It is worth keeping this broader
historical context in mind while enjoying the game's lighter treatment
of the period.
:::

## What is a decent score?

When the campaign starts, the board is minimal and a game is
relatively quick (less than an hour) --- and it is hard to tell what a
"good" score even looks like. For reference: in a normal game of
*TtR*, my usual score lands somewhere in the 100--150 range, so a
first game where the best player won with
\$`r max(scores$dollars[scores$year == 1865])` was a distinctly
foreign feeling. What we noticed is that as the campaign progressed,
the scores got *consistently* higher.

::: {.callout-note title="How tickets deplete" collapse="true"}
The more valuable tickets are sought after and eventually depleted:
one feature of the legacy system that keeps the ticket deck from
bloating is that whenever a company completes a ticket, it punches the
card --- some tickets allow one punch, others two --- and once full,
the ticket is retired to the Dead Letter Office. We return to *when*
the high-value tickets retired later in the report.
:::

```{r per-game}
#| fig-cap: >
#|   Dollars earned by each company throughout the campaign. The dashed
#|   grey line indicates the recommended dollar amount awarded to a
#|   player who missed that particular game, as specified by the
#|   end-of-campaign story deck card.
missing_scores <- dbGetQuery(con, "
  SELECT year, dollars FROM bank_slips WHERE player = 'missing'
") |>
  mutate(player = "(missing)")

pg_levels <- c(names(player_colors), "(missing)")
pg_colors <- c(player_colors, "(missing)" = "gray45")
pg_lines <- setNames(c(rep("solid", 5), "dashed"), pg_levels)

per_game <- bind_rows(
  mutate(scores, player = as.character(player)),
  missing_scores
) |>
  mutate(player = factor(player, levels = pg_levels))

ggplot(per_game, aes(year, dollars, color = player, linetype = player)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(per_game, player != "(missing)"), size = 2) +
  scale_color_manual(values = pg_colors) +
  scale_linetype_manual(values = pg_lines) +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Dollars per game",
    x = NULL, y = "dollars", color = NULL, linetype = NULL
  )
```

## The missing-player payout

First, a confession about how we handled absent players. The rulebook is
initially very vague about absent players: it merely says the campaign
can continue even when not everyone is present. We took that to mean a
missing player would simply receive *nothing* — hence the scheduling
nightmare and the long hiatus. Only at the end of the campaign does a
story deck card make up for those missed games, awarding a
**recommended dollar amount** for each game a player sat out. Rather
than attempting to recreate the absent player's result, the designers
simply award a fixed amount per game of the campaign — presumably
chosen to approximate what a typical company would have earned during
that game.

Because our campaign contains complete results for all five companies,
we can evaluate how well these recommendations match observed play.
Are they an unbiased estimate of an average player's earnings? Do they
systematically overestimate or underestimate actual performance? And
would following the rule have unfairly advantaged or disadvantaged a
player over the course of the campaign?

The remainder of this section compares the official "missing player"
values against the observed campaign results — effectively an
evaluation of one of the game's hidden balancing mechanisms.

### Game-level comparison

```{r missing-game-level}
official <- dbGetQuery(con, "
  SELECT year, dollars AS official FROM bank_slips WHERE player = 'missing'
")

obs <- scores |>
  group_by(year) |>
  summarise(
    mean = mean(dollars), median = median(dollars),
    sd = sd(dollars), .groups = "drop"
  ) |>
  left_join(official, by = "year") |>
  mutate(diff = mean - official)

knitr::kable(
  obs |> mutate(across(c(mean, median, sd, diff), ~ round(.x, 1))) |>
    select(year, official, mean, median, sd, `mean − official` = diff),
  caption = "Official missing-player amount vs observed earnings per game"
)
```

```{r missing-errors}
player_official <- scores |>
  left_join(official, by = "year") |>
  mutate(resid = dollars - official)

me   <- mean(player_official$resid)
mae  <- mean(abs(player_official$resid))
rmse <- sqrt(mean(player_official$resid^2))
```

Treating the guideline as a prediction of each company's earnings in
each game (60 observations), its **mean error is
\$`r round(me, 1)`** (positive = players actually earned more), with
a mean absolute error of \$`r round(mae, 1)` and RMSE of
\$`r round(rmse, 1)`. The bias is not constant, though — it grows
with the campaign economy, and the later games introduce outrageous
cash-cow rules: Sierra Madre's treasure hunt, and above all the
**Claims** that arrive with Cascadia, the final frontier (the CS-coded
scratch cards, each hiding \$98):

```{r missing-residuals}
#| fig-height: 4.5
#| fig-cap: >
#|   Residuals (actual − official) per game. The guideline tracks the
#|   early campaign well, then falls increasingly behind the inflating
#|   economy.
ggplot(player_official, aes(year, resid, color = player)) +
  geom_hline(yintercept = 0, color = "gray55", linetype = "dashed") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_color_manual(values = player_colors) +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Actual minus official, per game",
    x = NULL, y = "dollars above the guideline", color = NULL
  )
```

### Per-company performance

```{r missing-company}
#| fig-height: 4
#| fig-cap: >
#|   Distribution of (actual − official) across the 12 games. Boxes
#|   above zero mean the company out-earned the missing-player
#|   guideline.
ggplot(player_official, aes(player, resid, fill = player)) +
  geom_hline(yintercept = 0, color = "gray55", linetype = "dashed") +
  geom_boxplot(show.legend = FALSE, alpha = 0.8) +
  scale_fill_manual(values = player_colors) +
  labs(
    title = "Actual minus official, by company",
    x = NULL, y = "dollars above the guideline"
  )
```

```{r missing-tests}
tests <- player_official |>
  group_by(player) |>
  summarise(
    avg_dollars = round(mean(dollars), 1),
    avg_diff = round(mean(resid), 1),
    sd_diff = round(sd(resid), 1),
    p_wilcoxon = signif(
      wilcox.test(dollars, official, paired = TRUE, exact = FALSE)$p.value, 2
    ),
    .groups = "drop"
  )

knitr::kable(
  tests,
  col.names = c("company", "avg dollars", "avg − official",
                "sd", "Wilcoxon p"),
  caption = "Per-company deviation from the guideline
    (Wilcoxon signed-rank, paired, n = 12)"
)
```

### Interpretation

If the recommendations closely matched observed play, they would offer
a fair way to handle missed sessions. The data says they are
**increasingly stingy**: the guideline is roughly calibrated for the
first half of the campaign, but from 1883 onward the table's economy
inflates (bigger tickets, headquarters tolls, richer events) while the
guideline grows only linearly. Averaged over the whole campaign, four
of the five companies out-earned the guideline by \$13--19 per game;
only **Blue** essentially broke even against it (average difference
\$-1.4). With just 12 paired observations per company, none of the
Wilcoxon tests reach conventional significance (Green comes closest at
p = 0.074), so the per-company differences should be read as
descriptive. Still, the direction is consistent: skipping a
late-campaign game and taking the official payout would have been a
bad trade for everyone except perhaps Blue.

## Where do the points come from?

Bank slips break each game's score into components: coins in hand,
bonus for trains remaining, and completed-ticket value.

```{r components}
components <- scores |>
  select(player, year, coins_in_hand, trains_bonus, ticket_value) |>
  pivot_longer(-c(player, year), names_to = "component", values_to = "value") |>
  mutate(component = recode(component,
    coins_in_hand = "coins in hand",
    trains_bonus = "trains remaining bonus",
    ticket_value = "ticket value"
  ))

component_totals <- components |>
  group_by(player, component) |>
  summarise(total = sum(value, na.rm = TRUE), .groups = "drop")

ggplot(component_totals, aes(player, total, fill = player)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = player_colors) +
  facet_wrap(~component) +
  labs(
    title = "Campaign totals by scoring component",
    x = NULL, y = "dollars"
  )
```

```{r components-time}
#| fig-height: 6
ggplot(components, aes(year, value, color = player)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = player_colors) +
  facet_wrap(~component, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Scoring components over time",
    x = NULL, y = "dollars", color = NULL
  )
```

The trains-remaining bonus can be read backwards: the bank-slip
schedule pays $16/12/9/7/6 for 0–4 trains left and $4/2/0 for 5–7,
8–10, 11+ — so the bonus reveals how many trains each company had
left on the table:

```{r trains-remaining}
#| fig-height: 4.5
trains <- dbGetQuery(con, "SELECT * FROM v_trains_remaining") |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    trains_plot = case_when(
      is.na(trains_min)          ~ NA_real_,
      is.na(trains_max)          ~ 11,
      TRUE                       ~ (trains_min + trains_max) / 2
    )
  )

ggplot(trains, aes(year, trains_plot, color = player)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_color_manual(values = player_colors) +
  scale_x_continuous(breaks = unique(scores$year)) +
  scale_y_continuous(
    breaks = c(0:4, 6, 9, 11),
    labels = c("0", "1", "2", "3", "4", "5-7", "8-10", "11+")
  ) +
  labs(
    title = "Trains remaining at game end (derived from the bonus)",
    x = NULL, y = "trains remaining", color = NULL
  )
```

One bank slip needed an explanation: yellow's 1883 bonus of $28 is off
the schedule — a postcard effect that game paid $28 instead of the
usual $16 for 0 trains, so it plots as 0 trains remaining.

## Tickets

Punches per player — how many ticket completions each company recorded
over the whole campaign, split by normal tickets and postcard tickets
(single-punch tickets that retire immediately and earn a post card).

```{r punches}
punches <- dbGetQuery(con, "
  SELECT p.color AS player, c.capacity, SUM(p.punches) AS punches
  FROM ticket_punches p
  JOIN tickets t ON t.ticket_id = p.ticket_id
  JOIN card_ref_capacity c ON c.ref_id = t.ref_id AND c.color = p.color
  GROUP BY p.color, c.capacity
") |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    type = ifelse(capacity == 1, "postcard tickets", "normal tickets")
  )

ggplot(punches, aes(player, punches, fill = player)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = punches),
    vjust = -0.4, size = 3, color = "gray30"
  ) +
  scale_fill_manual(values = player_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.09))) +
  facet_wrap(~type) +
  labs(
    title = "Ticket punches per company",
    x = NULL, y = "punches"
  )
```

Ticket dollars actually collected (ticket value × punches):

```{r ticket-dollars}
ticket_dollars <- dbGetQuery(con, "
  SELECT p.color AS player, SUM(p.punches * t.value) AS dollars
  FROM ticket_punches p
  JOIN tickets t ON t.ticket_id = p.ticket_id
  GROUP BY p.color
") |>
  mutate(player = factor(player, levels = names(player_colors)))

ggplot(ticket_dollars, aes(player, dollars, fill = player)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = player_colors) +
  labs(
    title = "Total dollars from completed tickets",
    x = NULL, y = "dollars"
  )
```

## Claims

The endgame scratch cards. On partially-scratched cards the scratched
spots are exactly the claims collected during play; fully-scratched cards
were opened out of curiosity afterwards, so their in-play earnings are
unknown.

```{r claims}
claims <- dbGetQuery(con, "SELECT * FROM v_claim_earnings")
knitr::kable(
  claims |>
    mutate(
      earned = ifelse(earnings_known == 1, earned_dollars, "unknown (fully scratched)"),
      owner = coalesce(owner, note, "?")
    ) |>
    select(card = card_code, town, owner, earned),
  caption = "Claim card earnings (all cards hide the same $98 total but in random ordering)"
)
```

## Company shares

Sixty shares across six railway companies (including the neutral white
railway), ten per company. Most shares in a company ranks first (at
least one share required to qualify; ties broken by the lowest-numbered
share held), paying $20/15/10/5 down the ranks.

```{r shares}
#| fig-height: 5
company_colors <- c(player_colors, white = "#9e9e9e", total = "#555555")

shares <- dbGetQuery(con, "SELECT * FROM shares")

share_totals <- shares |>
  group_by(player) |>
  summarise(qty = sum(qty), .groups = "drop") |>
  mutate(company = "total")

shares_plot <- bind_rows(shares, share_totals) |>
  group_by(company) |>
  mutate(alpha_val = ifelse(qty == 0, 0, 0.15 + 0.85 * qty / max(qty))) |>
  ungroup() |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    company = factor(company,
      levels = rev(c("black", "blue", "green", "yellow", "red", "white", "total"))
    )
  )

ggplot(shares_plot, aes(player, company)) +
  geom_tile(aes(fill = company, alpha = alpha_val),
    color = "white", linewidth = 1.2
  ) +
  geom_text(
    aes(label = ifelse(qty > 0, qty, ""),
        color = alpha_val > 0.6),
    size = 3.6, fontface = "bold"
  ) +
  scale_fill_manual(values = company_colors, guide = "none") +
  scale_alpha_identity() +
  scale_color_manual(values = c(`TRUE` = "white", `FALSE` = "gray25"),
    guide = "none"
  ) +
  coord_fixed(ratio = 0.6) +
  labs(
    title = "Share holdings",
    caption = "Each company has 10 shares.",
    x = "players", y = "company"
  )
```

```{r share-payouts}
#| fig-height: 3.5
payouts <- dbGetQuery(con, "SELECT * FROM share_payouts") |>
  mutate(player = factor(player, levels = names(player_colors)))

ggplot(payouts, aes(player, dollars, fill = player)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = dollars), vjust = -0.4, size = 3, color = "gray30") +
  scale_fill_manual(values = player_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  labs(
    title = "Share dollars ($295 paid out in total)",
    x = NULL, y = "dollars"
  )
```

Green's diversified portfolio (22 of 60 shares, holding stakes in every
company) pays $110 — more than a third of the entire pool.

## The board

The game board as a network: 200 tracks connecting 63 cities across 12
jigsaw sections. Fictional cities (Nuevos Angeles, Cemetery City,
Pacific Haven, Baja) sit at plausible approximate positions; dashed
gray = unbuilt track beds. The cities are colored by the sticker
locations of *our* campaign — your board may vary.

```{r board}
#| fig-height: 7
coords <- dbGetQuery(con, "
  SELECT cc.*, c.company, c.port
  FROM city_coords cc JOIN cities c ON c.city = cc.city
") |>
  mutate(
    port_type = ifelse(coalesce(port, 0) == 1, "port", "inland"),
    company = ifelse(is.na(company), NA, company)
  )
routes <- dbGetQuery(con, "
  SELECT r.*, a.lat AS lat_a, a.lon AS lon_a, b.lat AS lat_b, b.lon AS lon_b
  FROM routes r
  JOIN city_coords a ON a.city = r.city_a
  JOIN city_coords b ON b.city = r.city_b
")

track_palette <- c(
  black = "#1a1a1a", blue = "#1f77c4", green = "#2a9d3a",
  yellow = "#e6b400", red = "#d62728", white = "#c8c8c8",
  gray = "#8a8a8a"
)

sticker_colors <- c(
  player_colors,
  "Mama O'Connel" = "#8B5A2B",   # brown
  big = "#9467bd"                 # purple
)

ggplot() +
  geom_segment(
    data = filter(routes, is.na(track_color)),
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray70", linetype = "dashed", linewidth = 0.5
  ) +
  geom_segment(
    data = filter(routes, !is.na(track_color)),
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b,
        color = track_color, linewidth = length),
    alpha = 0.55
  ) +
  geom_point(
    data = coords,
    aes(lon, lat, fill = company, shape = port_type),
    size = 2.6, color = "gray20", stroke = 0.5
  ) +
  geom_text(
    data = coords, aes(lon, lat, label = city),
    size = 2.4, vjust = -1.0, check_overlap = TRUE
  ) +
  scale_color_manual(values = track_palette, guide = "none") +
  scale_fill_manual(
    values = sticker_colors, na.value = NA,
    name = "company sticker",
    guide = guide_legend(override.aes = list(shape = 21, size = 3))
  ) +
  scale_shape_manual(
    values = c(inland = 21, port = 24),
    name = NULL,
    guide = guide_legend(override.aes = list(size = 3))
  ) +
  scale_linewidth_continuous(range = c(0.3, 1.8), guide = "none") +
  coord_quickmap() +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(title = "The board: tracks as printed, cities by company sticker (triangles = ports)")
```

### Opening the frontiers

How the map grows (per the official rules): at the end of a game, the
winner chooses which jigsaw piece to add next — its cities and tickets
come into play from the *next* game on. The dead letter office
remembers the order: frontier cards sit between dated story markers in
the pile, giving both the sequence and (with the one-frontier-per-game
constraint) the year each section became playable. Drag the slider (or
press play) to watch the board unfold from the initial East Coast start
area; cities are colored by jigsaw section, and tracks appear once both
their endpoints are unlocked.

```{r frontier-unlock}
#| fig-height: 6.5
frontier_map <- c(
  "great plains" = "Great Plains", "florida" = "Florida",
  "sierra madre" = "Sierra Madre", "haunted waste" = "Haunted Wastes",
  "badlands" = "Badlands", "california" = "California",
  "open range" = "Open Range", "cascadia" = "Cascadia"
)

unlocks <- dbGetQuery(con, "
  SELECT s.id, s.name, w.latest_year
  FROM story_cards s
  JOIN v_retirement_window w ON w.run_id = s.id
") |>
  mutate(key = tolower(name)) |>
  filter(key %in% names(frontier_map)) |>
  arrange(desc(id)) |>   # deeper in the pile = chosen earlier
  transmute(section = unname(frontier_map[key]), year = latest_year) |>
  # a frontier is chosen at the end of one game and playable from the
  # next, so in-play years must strictly increase (one per game); this
  # also corrects two markers that sit scrambled in the pile
  mutate(year = Reduce(
    function(prev, cur) max(cur, prev + 3),
    year, accumulate = TRUE
  ))

initial_sections <- c("Northeast", "Atlantic Coast", "The South", "Appalachia")

city_sections <- dbGetQuery(con, "
  SELECT c.city, c.section, cc.lat, cc.lon FROM cities c
  JOIN city_coords cc ON cc.city = c.city
")

# cumulative frames: each step adds one frontier's sections
acc <- initial_sections
frame_defs <- list(tibble(frame = "1865 · Initial", section = acc))
for (i in seq_len(nrow(unlocks))) {
  acc <- c(acc, unlocks$section[i])
  frame_defs[[i + 1]] <- tibble(
    frame = paste0(unlocks$year[i], " · ", unlocks$section[i]),
    section = acc
  )
}
frames_df <- bind_rows(frame_defs)

cities_anim <- frames_df |>
  inner_join(city_sections, by = "section",
    relationship = "many-to-many"
  ) |>
  mutate(tooltip = paste0("<b>", city, "</b><br>", section))

routes_anim <- frames_df |>
  distinct(frame) |>
  cross_join(routes) |>
  semi_join(
    cities_anim |> select(frame, city),
    by = c("frame", "city_a" = "city")
  ) |>
  semi_join(
    cities_anim |> select(frame, city),
    by = c("frame", "city_b" = "city")
  )

p_unlock <- ggplot() +
  geom_segment(
    data = routes_anim,
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b, frame = frame),
    color = "gray80", linewidth = 0.4
  ) +
  geom_point(
    data = cities_anim,
    aes(lon, lat, color = section, frame = frame, text = tooltip),
    size = 2.4
  ) +
  coord_fixed(ratio = 1.25) +
  theme_void() +
  labs(title = "How the board unfolded", color = "jigsaw section")

ggplotly(p_unlock, tooltip = "text") |>
  animation_opts(frame = 1200, transition = 300, redraw = FALSE) |>
  animation_slider(currentvalue = list(prefix = "unlocked: ", font = list(size = 14))) |>
  config(displayModeBar = FALSE)
```

### Hotspots

Which cities did completed tickets actually connect? Each punch on a
ticket counts once for each of its two endpoint cities. **Hover a
city** for its name, punch count, and most frequent visitor — the
company that punched the most tickets through it.

```{r hotspots}
#| fig-height: 6.5
city_players <- dbGetQuery(con, "
  WITH endpoint_punches AS (
    SELECT t.from_city AS city, p.color AS player, p.punches FROM tickets t
    JOIN ticket_punches p ON p.ticket_id = t.ticket_id
    UNION ALL
    SELECT t.to_city, p.color, p.punches FROM tickets t
    JOIN ticket_punches p ON p.ticket_id = t.ticket_id
  )
  SELECT city, player, SUM(punches) AS punches
  FROM endpoint_punches GROUP BY city, player
")

top_visitors <- city_players |>
  filter(punches > 0) |>
  group_by(city) |>
  filter(punches == max(punches)) |>
  summarise(
    top_label = paste0(paste(player, collapse = " & "),
                       " (", first(punches), ")"),
    .groups = "drop"
  )

hot <- city_players |>
  group_by(city) |>
  summarise(punches = sum(punches), .groups = "drop") |>
  left_join(top_visitors, by = "city") |>
  inner_join(coords, by = "city") |>
  mutate(tooltip = paste0(
    "<b>", city, "</b><br>punches: ", punches,
    "<br>most frequent visitor: ", coalesce(top_label, "none")
  ))

p_hot <- ggplot() +
  geom_segment(
    data = routes,
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray85", linewidth = 0.4
  ) +
  geom_point(
    data = hot,
    aes(lon, lat, size = punches, color = punches, text = tooltip),
    alpha = 0.8
  ) +
  scale_size_continuous(range = c(1, 9)) +
  scale_color_viridis_c(option = "inferno", direction = -1) +
  coord_fixed(ratio = 1.25) +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "Ticket hotspots: endpoint cities weighted by punches")

ggplotly(p_hot, tooltip = "text") |>
  config(displayModeBar = FALSE)
```

### When was the network built?

Retired tickets sit between dated story markers in the dead letter
office, which bounds when they were completed. Using the midpoint of
each ticket's retirement window as a rough timestamp:

```{r eras}
#| fig-height: 5
era_tickets <- dbGetQuery(con, "
  SELECT t.from_city, t.to_city,
         (COALESCE(w.earliest_year, 1865) + COALESCE(w.latest_year, 1901)) / 2 AS mid_year,
         SUM(p.punches) AS punches
  FROM tickets t
  JOIN v_retirement_window w ON w.run_id = t.run_id
  JOIN ticket_punches p ON p.ticket_id = t.ticket_id
  WHERE t.active = 0
  GROUP BY t.ticket_id
  HAVING punches > 0
") |>
  mutate(era = cut(mid_year,
    breaks = c(1864, 1874, 1886, 1901),
    labels = c("early (1865-1874)", "mid (1877-1886)", "late (1889-1901)")
  )) |>
  inner_join(coords, by = c("from_city" = "city")) |>
  rename(lat_a = lat, lon_a = lon) |>
  inner_join(coords, by = c("to_city" = "city")) |>
  rename(lat_b = lat, lon_b = lon)

ggplot(era_tickets) +
  geom_segment(
    data = routes,
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray88", linewidth = 0.3
  ) +
  geom_segment(
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b,
        linewidth = punches),
    color = "#c1440e", alpha = 0.5
  ) +
  scale_linewidth_continuous(range = c(0.3, 1.6), guide = "none") +
  coord_quickmap() +
  facet_wrap(~era) +
  theme_void() +
  labs(
    title = "Completed ticket connections by campaign era",
    subtitle = "Straight lines connect ticket endpoints (not actual track paths)"
  )
```

## Next up

- Route the ticket connections along actual board tracks (shortest path
  over the `routes` network) instead of straight lines.
- Claim / postcard geography: where the postcard tickets cluster.

```{r close}
#| include: false
dbDisconnect(con)
```
