---
title: "Ticket to Ride Legacy — Data Dive"
subtitle: "One campaign, 13 games, 5 companies — reconstructed from the dead letter office"
---

```{r setup}
#| include: false
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)

con <- dbConnect(SQLite(), "ttr.db")

player_colors <- c(
  black  = "#1a1a1a",
  blue   = "#1f77c4",
  green  = "#2a9d3a",
  yellow = "#e6b400",
  red    = "#d62728"
)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold")
    )
)

scores <- dbGetQuery(con, "SELECT * FROM v_scores") |>
  mutate(player = factor(player, levels = names(player_colors)))

players <- dbGetQuery(con, "SELECT * FROM players")
```

## Score progression

Dollars earned per game, per company. Each game is one year of the
campaign, 1865–1898. The dashed gray line is the "(missing)" baseline
recorded on the bank-slip pad.

```{r per-game}
missing_scores <- dbGetQuery(con, "
  SELECT year, dollars FROM bank_slips WHERE player = 'missing'
")

ggplot(scores, aes(year, dollars, color = player)) +
  geom_line(
    data = mutate(missing_scores, player = "(missing)"),
    aes(year, dollars), linetype = "dashed", color = "gray45",
    linewidth = 0.8, inherit.aes = FALSE
  ) +
  annotate("text",
    x = max(missing_scores$year), y = max(missing_scores$dollars),
    label = "(missing)", hjust = 1, vjust = -0.8,
    size = 3.2, color = "gray45"
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = player_colors) +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Dollars per game",
    x = NULL, y = "dollars", color = NULL
  )
```

The campaign total is what decides the winner — the running sum of all
game scores, plus what was found in each company's vault after the
campaign (post cards and share certificates; shown as the final "vault"
step after 1898):

```{r cumulative}
vault <- dbGetQuery(con, "
  SELECT vault AS player, SUM(CAST(worth AS INTEGER)) AS vault_dollars
  FROM post_office
  WHERE vault IS NOT NULL AND worth NOT GLOB '*[a-z]*'
  GROUP BY vault
")

cum_scores <- scores |>
  arrange(player, year) |>
  group_by(player) |>
  mutate(total = cumsum(dollars)) |>
  ungroup()

vault_step <- cum_scores |>
  filter(year == max(year)) |>
  left_join(vault, by = "player") |>
  mutate(
    year = 1901,
    total = total + coalesce(vault_dollars, 0)
  )

cum_with_vault <- bind_rows(cum_scores, vault_step)

ggplot(cum_with_vault, aes(year, total, color = player)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = player_colors) +
  scale_x_continuous(
    breaks = c(unique(scores$year), 1901),
    labels = c(unique(scores$year), "vault")
  ) +
  labs(
    title = "Campaign total (cumulative dollars, vault additions last)",
    x = NULL, y = "cumulative dollars", color = NULL
  )
```

Blue's vault also held a *piggyback* post card (not a dollar value, so
not included in the step).

### Did the ordering change?

Rank 1 = leading. First the rank within each individual game, then the
rank on cumulative campaign total — the race that matters.

```{r rank-helpers}
#| include: false
# sigmoid interpolation between rank points, sankey/bump style
bump_curves <- function(df, n = 40) {
  df <- arrange(df, player, year)
  do.call(rbind, lapply(split(df, df$player), function(d) {
    segs <- lapply(seq_len(nrow(d) - 1), function(i) {
      t <- seq(0, 1, length.out = n)
      s <- 1 / (1 + exp(-12 * (t - 0.5)))   # sigmoid easing
      data.frame(
        player = d$player[i],
        x = d$year[i] + t * (d$year[i + 1] - d$year[i]),
        y = d$rank[i] + s * (d$rank[i + 1] - d$rank[i])
      )
    })
    do.call(rbind, segs)
  }))
}

plot_bump <- function(rank_df, title) {
  curves <- bump_curves(rank_df)
  ends <- filter(rank_df, year == max(year))
  ggplot() +
    geom_line(
      data = curves, aes(x, y, color = player, group = player),
      linewidth = 2.4, alpha = 0.85, lineend = "round"
    ) +
    geom_point(data = rank_df, aes(year, rank, color = player), size = 3.2) +
    geom_text(
      data = ends,
      aes(year, rank, label = player, color = player),
      hjust = -0.35, size = 3.4, fontface = "bold", show.legend = FALSE
    ) +
    scale_color_manual(values = player_colors, guide = "none") +
    scale_y_reverse(breaks = 1:5) +
    scale_x_continuous(
      breaks = unique(rank_df$year),
      expand = expansion(mult = c(0.02, 0.09))
    ) +
    labs(title = title, x = NULL, y = "rank")
}
```

```{r ranks-game}
#| fig-height: 4
ranks <- cum_scores |>
  group_by(year) |>
  mutate(
    game_rank = rank(-dollars, ties.method = "min"),
    campaign_rank = rank(-total, ties.method = "min")
  ) |>
  ungroup()

plot_bump(
  transmute(ranks, player, year, rank = game_rank),
  "Rank within each game"
)
```

```{r ranks-campaign}
#| fig-height: 4
plot_bump(
  transmute(ranks, player, year, rank = campaign_rank),
  "Rank on campaign total"
)
```

## Where do the points come from?

Bank slips break each game's score into components: coins in hand,
bonus for trains remaining, and completed-ticket value.

```{r components}
components <- scores |>
  select(player, year, coins_in_hand, trains_bonus, ticket_value) |>
  pivot_longer(-c(player, year), names_to = "component", values_to = "value") |>
  mutate(component = recode(component,
    coins_in_hand = "coins in hand",
    trains_bonus = "trains remaining bonus",
    ticket_value = "ticket value"
  ))

component_totals <- components |>
  group_by(player, component) |>
  summarise(total = sum(value, na.rm = TRUE), .groups = "drop")

ggplot(component_totals, aes(player, total, fill = player)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = player_colors) +
  facet_wrap(~component) +
  labs(
    title = "Campaign totals by scoring component",
    x = NULL, y = "dollars"
  )
```

```{r components-time}
#| fig-height: 6
ggplot(components, aes(year, value, color = player)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = player_colors) +
  facet_wrap(~component, ncol = 1) +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Scoring components over time",
    x = NULL, y = "dollars", color = NULL
  )
```

## Tickets

Punches per player — how many ticket completions each company recorded
over the whole campaign, split by normal tickets and postcard tickets
(single-punch tickets that retire immediately and earn a post card).

```{r punches}
punches <- dbGetQuery(con, "
  SELECT p.color AS player, c.capacity, SUM(p.punches) AS punches
  FROM ticket_punches p
  JOIN tickets t ON t.ticket_id = p.ticket_id
  JOIN card_ref_capacity c ON c.ref_id = t.ref_id AND c.color = p.color
  GROUP BY p.color, c.capacity
") |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    type = ifelse(capacity == 1, "postcard tickets", "normal tickets")
  )

ggplot(punches, aes(player, punches, fill = player)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = punches),
    vjust = -0.4, size = 3, color = "gray30"
  ) +
  scale_fill_manual(values = player_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.09))) +
  facet_wrap(~type) +
  labs(
    title = "Ticket punches per company",
    x = NULL, y = "punches"
  )
```

Ticket dollars actually collected (ticket value × punches):

```{r ticket-dollars}
ticket_dollars <- dbGetQuery(con, "
  SELECT p.color AS player, SUM(p.punches * t.value) AS dollars
  FROM ticket_punches p
  JOIN tickets t ON t.ticket_id = p.ticket_id
  GROUP BY p.color
") |>
  mutate(player = factor(player, levels = names(player_colors)))

ggplot(ticket_dollars, aes(player, dollars, fill = player)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = player_colors) +
  labs(
    title = "Total dollars from completed tickets",
    x = NULL, y = "dollars"
  )
```

## Claims

The endgame scratch cards. On partially-scratched cards the scratched
spots are exactly the claims collected during play; fully-scratched cards
were opened out of curiosity afterwards, so their in-play earnings are
unknown.

```{r claims}
claims <- dbGetQuery(con, "SELECT * FROM v_claim_earnings")
knitr::kable(
  claims |>
    mutate(
      earned = ifelse(earnings_known == 1, earned_dollars, "unknown (fully scratched)"),
      owner = coalesce(owner, note, "?")
    ) |>
    select(card = card_code, town, owner, earned),
  caption = "Claim card earnings (all cards hide the same 98$ total but in random ordering)"
)
```

## The board

The game board as a network: 200 tracks connecting 63 cities across 12
jigsaw sections. Fictional cities (Nuevos Angeles, Cemetery City,
Pacific Haven, Baja) sit at plausible approximate positions; dashed
gray = unbuilt track beds.

```{r board}
#| fig-height: 7
coords <- dbGetQuery(con, "SELECT * FROM city_coords")
routes <- dbGetQuery(con, "
  SELECT r.*, a.lat AS lat_a, a.lon AS lon_a, b.lat AS lat_b, b.lon AS lon_b
  FROM routes r
  JOIN city_coords a ON a.city = r.city_a
  JOIN city_coords b ON b.city = r.city_b
")

track_palette <- c(
  black = "#1a1a1a", blue = "#1f77c4", green = "#2a9d3a",
  yellow = "#e6b400", red = "#d62728", white = "#c8c8c8",
  gray = "#8a8a8a"
)

ggplot() +
  geom_segment(
    data = filter(routes, is.na(track_color)),
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray70", linetype = "dashed", linewidth = 0.5
  ) +
  geom_segment(
    data = filter(routes, !is.na(track_color)),
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b,
        color = track_color, linewidth = length),
    alpha = 0.55
  ) +
  geom_point(data = coords, aes(lon, lat), size = 1.6, color = "gray20") +
  geom_text(
    data = coords, aes(lon, lat, label = city),
    size = 2.4, vjust = -0.9, check_overlap = TRUE
  ) +
  scale_color_manual(values = track_palette, guide = "none") +
  scale_linewidth_continuous(range = c(0.3, 1.8), guide = "none") +
  coord_quickmap() +
  theme_void() +
  labs(title = "The board: every track, colored as printed")
```

### Hotspots

Which cities did completed tickets actually connect? Each punch on a
ticket counts once for each of its two endpoint cities.

```{r hotspots}
#| fig-height: 7
hot <- dbGetQuery(con, "
  WITH endpoint_punches AS (
    SELECT t.from_city AS city, p.punches FROM tickets t
    JOIN ticket_punches p ON p.ticket_id = t.ticket_id
    UNION ALL
    SELECT t.to_city, p.punches FROM tickets t
    JOIN ticket_punches p ON p.ticket_id = t.ticket_id
  )
  SELECT city, SUM(punches) AS punches
  FROM endpoint_punches GROUP BY city
") |>
  inner_join(coords, by = "city")

ggplot() +
  geom_segment(
    data = routes,
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray85", linewidth = 0.4
  ) +
  geom_point(
    data = hot,
    aes(lon, lat, size = punches, color = punches),
    alpha = 0.8
  ) +
  geom_text(
    data = slice_max(hot, punches, n = 12),
    aes(lon, lat, label = paste0(city, " (", punches, ")")),
    size = 2.8, vjust = -1.1, fontface = "bold", check_overlap = TRUE
  ) +
  scale_size_continuous(range = c(1, 10)) +
  scale_color_viridis_c(option = "inferno", direction = -1) +
  coord_quickmap() +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "Ticket hotspots: endpoint cities weighted by punches")
```

### When was the network built?

Retired tickets sit between dated story markers in the dead letter
office, which bounds when they were completed. Using the midpoint of
each ticket's retirement window as a rough timestamp:

```{r eras}
#| fig-height: 5
era_tickets <- dbGetQuery(con, "
  SELECT t.from_city, t.to_city,
         (COALESCE(w.earliest_year, 1865) + COALESCE(w.latest_year, 1901)) / 2 AS mid_year,
         SUM(p.punches) AS punches
  FROM tickets t
  JOIN v_retirement_window w ON w.run_id = t.run_id
  JOIN ticket_punches p ON p.ticket_id = t.ticket_id
  WHERE t.active = 0
  GROUP BY t.ticket_id
  HAVING punches > 0
") |>
  mutate(era = cut(mid_year,
    breaks = c(1864, 1874, 1886, 1901),
    labels = c("early (1865-1874)", "mid (1877-1886)", "late (1889-1901)")
  )) |>
  inner_join(coords, by = c("from_city" = "city")) |>
  rename(lat_a = lat, lon_a = lon) |>
  inner_join(coords, by = c("to_city" = "city")) |>
  rename(lat_b = lat, lon_b = lon)

ggplot(era_tickets) +
  geom_segment(
    data = routes,
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray88", linewidth = 0.3
  ) +
  geom_segment(
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b,
        linewidth = punches),
    color = "#c1440e", alpha = 0.5
  ) +
  scale_linewidth_continuous(range = c(0.3, 1.6), guide = "none") +
  coord_quickmap() +
  facet_wrap(~era) +
  theme_void() +
  labs(
    title = "Completed ticket connections by campaign era",
    subtitle = "Straight lines connect ticket endpoints (not actual track paths)"
  )
```

## Next up

- Route the ticket connections along actual board tracks (shortest path
  over the `routes` network) instead of straight lines.
- Claim / postcard geography: where the postcard tickets cluster.

```{r close}
#| include: false
dbDisconnect(con)
```
