---
title: "Ticket to Ride Legacy — Data Dive"
subtitle: "One campaign, 12 plays, 5 companies — reconstructed from the dead letter office"
---

```{r setup}
#| include: false
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)

con <- dbConnect(SQLite(), "ttr.db")

player_colors <- c(
  black  = "#1a1a1a",
  blue   = "#1f77c4",
  green  = "#2a9d3a",
  yellow = "#e6b400",
  red    = "#d62728"
)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold")
    )
)

scores <- dbGetQuery(con, "SELECT * FROM v_scores") |>
  mutate(player = factor(player, levels = names(player_colors)))

players <- dbGetQuery(con, "SELECT * FROM players")
```

## What on earth is Ticket to Ride Legacy?

[*Ticket to Ride*](https://boardgamegeek.com/boardgame/9209/ticket-to-ride)
(TtR) is Alan R. Moon's famous gateway board game — the set-collecting,
network-building train game; most people where I come from know the
[*Europe*](https://boardgamegeek.com/boardgame/14996/ticket-to-ride-europe)
or
[*Nordic*](https://boardgamegeek.com/boardgame/31627/ticket-to-ride-nordic-countries)
variety. Each TtR variation is basically draw colored train cards,
claim routes between cities, complete tickets for points — with
various nuances for additional scoring (like longest road).
[*Ticket to Ride Legacy: Legends of the West*](https://boardgamegeek.com/boardgame/390092/ticket-to-ride-legacy-legends-of-the-west)
(2023) gives that familiar game the **legacy treatment** pioneered by
Rob Daviau and Matt Leacock in *Pandemic Legacy* (seasons
[1](https://boardgamegeek.com/boardgame/161936/pandemic-legacy-season-1),
[2](https://boardgamegeek.com/boardgame/221107/pandemic-legacy-season-2)
and [0](https://boardgamegeek.com/boardgame/314040/pandemic-legacy-season-0)
— yes, in that order): a 12-game campaign where the results of each
play permanently change the next — the board grows piece by piece,
cards are punched and stickered, and unlike many legacy systems
nothing gets torn up: it might be altered, but eventually it gets
retired for good into the dead letter office, while the story unfolds
as you go.

::: {.callout-warning title="Spoiler warning"}
This report is one long spoiler. It shows the full board, every
frontier, hidden card values, postcards, and campaign-end scoring. If
you ever plan to play *Ticket to Ride Legacy: Legends of the West*
unspoiled, close this tab now and come back in twelve games.
:::

## Our campaign

Our campaign was played by five players. Throughout this report, they
are referred to by their player colours — **Green**, **Black**,
**Red**, **Yellow**, and **Blue**. As each player controls a railroad
company throughout the campaign, the terms *player* and *company* are
used interchangeably.

The game itself was a Christmas present from Green and Black to
Black's sisters, Red and Yellow, and Yellow's husband, Blue. Everyone
at the table was already well versed in *Ticket to Ride* —
particularly *Ticket to Ride: Europe* — and enjoyed set-collection
train games. Only Green and Black had previous experience with legacy
games, having completed *Pandemic Legacy* Seasons 1 and 2 (with
Season 0 still waiting on the shelf of opportunity). As those are
cooperative games, where players work together against the game,
*Legends of the West* was everyone's **first competitive legacy
campaign**.

The campaign consists of twelve games but, in our case, spanned
roughly **two and a half years**. Although the game was received as a
Christmas gift in 2023 and the first session took place in February
2024, the campaign included a lengthy hiatus from late 2024 until
April 2026. In the meantime, life happened: pregnancies, the arrival
of two new family members in 2025, and the inevitable challenge of
finding time to get everyone back around the table.

After the campaign concluded, the complete game state was
painstakingly **reconstructed by hand**. Every remaining card in the
Storeroom, Post Office, Company Vaults, and the entire Dead Letter
Office pile was catalogued card by card and entered into the database
that forms the basis of the analyses presented in this report.

## Score progression

Unlike regular Ticket to Ride, the game isn't about points exactly,
but **dollars**. You always start a game with $5 (sometimes more,
depending on your employee cards — asymmetric player abilities dealt
for that particular game). An event deck can gain you more money (read:
points) or lose you some. Later in the campaign your company can put
up **headquarters** in cities (though not in big cities), and every
other player laying a route into that city owes you $2. And there is
an overarching villain in TtR Legacy — **Mama O'Connell** — whom you
might have to pay instead, meaning the money goes back to the bank.

Dollars earned per game, per company. Each play is one game of the
campaign, and the campaign calendar jumps three years per game:
`1865:3:1898`. The dashed gray line is the suggested dollars for
a player missing in action that game — a baseline value.

::: {.callout-note title="Why 1865–1898?" collapse="true"}
The campaign rides the great American railroad boom: it opens in 1865,
the year the Civil War ended and the race west began in earnest (the
transcontinental golden spike followed in 1869), and closes as the
1890s wind down — the decade the frontier was declared closed. The
rulebook leans on the era's long-lasting tropes — expansionism, the
Wild West, the great outdoors — and its designers' note is upfront
that the real expansion came at a cost the game's lighter theme
doesn't depict: the dispossession of Native American tribes and the
exploitation of railroad laborers. Worth reading about alongside the
game.
:::

```{r per-game}
missing_scores <- dbGetQuery(con, "
  SELECT year, dollars FROM bank_slips WHERE player = 'missing'
") |>
  mutate(player = "(missing)")

pg_levels <- c(names(player_colors), "(missing)")
pg_colors <- c(player_colors, "(missing)" = "gray45")
pg_lines <- setNames(c(rep("solid", 5), "dashed"), pg_levels)

per_game <- bind_rows(
  mutate(scores, player = as.character(player)),
  missing_scores
) |>
  mutate(player = factor(player, levels = pg_levels))

ggplot(per_game, aes(year, dollars, color = player, linetype = player)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(per_game, player != "(missing)"), size = 2) +
  scale_color_manual(values = pg_colors) +
  scale_linetype_manual(values = pg_lines) +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Dollars per game",
    x = NULL, y = "dollars", color = NULL, linetype = NULL
  )
```

The campaign total is what decides the winner — the running sum of the
bank slips through 1898, then each post-campaign addition stepped in
one at a time: post cards (from the vaults), circus flyer, shares, and
finally the timetable. The number at the right is each company's final
score.

```{r cumulative}
#| fig-height: 6
cum_scores <- scores |>
  arrange(player, year) |>
  group_by(player) |>
  mutate(total = cumsum(dollars)) |>
  ungroup()

component_order <- c("postcards", "circus", "shares", "timetable")
component_x <- setNames(1901 + 3 * seq_along(component_order), component_order)

additions_wide <- dbGetQuery(con, "
  SELECT player, component, dollars FROM final_components
  WHERE component <> 'bankslips'
")

steps <- cum_scores |>
  filter(year == max(year)) |>
  select(player, total)
step_rows <- list()
for (comp in component_order) {
  steps <- steps |>
    left_join(filter(additions_wide, component == comp) |>
        select(player, add = dollars), by = "player") |>
    mutate(total = total + coalesce(add, 0)) |>
    select(-add)
  step_rows[[comp]] <- mutate(steps, year = component_x[[comp]])
}

cum_full <- bind_rows(
  select(cum_scores, player, year, total),
  bind_rows(step_rows)
) |>
  mutate(player = factor(player, levels = names(player_colors)))

totals_lab <- cum_full |>
  filter(year == max(year)) |>
  arrange(total) |>
  mutate(
    lab_y = total,
    # push overlapping labels apart (min 45 dollar-units of separation)
    lab_y = Reduce(
      function(prev, cur) max(cur, prev + 45),
      lab_y, accumulate = TRUE
    )
  )

ggplot(cum_full, aes(year, total, color = player)) +
  annotate("segment",
    x = 1865, xend = 1898, y = -95, yend = -95,
    color = "gray55", linewidth = 0.5
  ) +
  annotate("text",
    x = (1865 + 1898) / 2, y = -160, label = "bank slips",
    color = "gray45", size = 3.2
  ) +
  annotate("segment",
    x = min(component_x), xend = max(component_x), y = -95, yend = -95,
    color = "gray55", linewidth = 0.5
  ) +
  annotate("text",
    x = mean(range(component_x)), y = -160, label = "vault",
    color = "gray45", size = 3.2
  ) +
  geom_vline(xintercept = 1901, color = "gray80", linetype = "dotted") +
  annotate("text",
    x = 1900, y = 650, label = "post campaign",
    angle = 90, color = "gray55", size = 3
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_text(
    data = totals_lab,
    aes(y = lab_y, label = paste0("$", total)),
    hjust = -0.2, size = 3.3, fontface = "bold", show.legend = FALSE
  ) +
  scale_color_manual(values = player_colors) +
  scale_x_continuous(
    breaks = c(unique(scores$year), component_x),
    labels = c(unique(scores$year), "post cards", "circus flyer",
               "shares", "timetable"),
    expand = expansion(mult = c(0.02, 0.06))
  ) +
  coord_cartesian(clip = "off") +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    plot.margin = margin(10, 20, 26, 10)
  ) +
  labs(
    title = "Campaign total (bank slips, then each addition stepped in)",
    x = NULL, y = "cumulative dollars", color = NULL
  )
```

The post-campaign vault additions nearly rewrote the podium: green's $259 of
additions ($110 shares, $71 circus, $30 timetable, $48 post cards)
closed the gap to black from $66 to just **$18** — final score $1363 to
$1345.

```{r final-additions}
#| fig-height: 4
additions <- dbGetQuery(con, "
  SELECT player, component, dollars
  FROM final_components
  WHERE component <> 'bankslips'
") |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    component = factor(component,
      levels = c("shares", "circus", "timetable", "postcards")
    )
  )

ggplot(additions, aes(player, dollars, fill = component)) +
  geom_col(color = "white", linewidth = 0.3) +
  scale_fill_manual(values = c(
    shares = "#4477aa", circus = "#ee6677",
    timetable = "#228833", postcards = "#ccbb44"
  )) +
  labs(
    title = "Post-campaign vault additions by component",
    x = NULL, y = "dollars", fill = NULL
  )
```

### Did the ordering change?

Rank 1 = leading. First the rank within each individual game — extended
past 1898 with the rank within each vault addition (who earned the most
post cards, circus flyer, shares, timetable) and the final total. Then
the rank on the running campaign total as each of those steps in — the
race that matters.

```{r rank-helpers}
#| include: false
# sigmoid interpolation between rank points, sankey/bump style
bump_curves <- function(df, n = 40) {
  df <- arrange(df, player, year)
  do.call(rbind, lapply(split(df, df$player), function(d) {
    segs <- lapply(seq_len(nrow(d) - 1), function(i) {
      t <- seq(0, 1, length.out = n)
      s <- 1 / (1 + exp(-12 * (t - 0.5)))   # sigmoid easing
      data.frame(
        player = d$player[i],
        x = d$year[i] + t * (d$year[i + 1] - d$year[i]),
        y = d$rank[i] + s * (d$rank[i + 1] - d$rank[i])
      )
    })
    do.call(rbind, segs)
  }))
}

plot_bump <- function(rank_df, title, x_breaks, x_labels) {
  curves <- bump_curves(rank_df)
  ends <- filter(rank_df, year == max(year))
  ggplot() +
    geom_vline(xintercept = 1901, color = "gray80", linetype = "dotted") +
    geom_line(
      data = curves, aes(x, y, color = player, group = player),
      linewidth = 2.4, alpha = 0.85, lineend = "round"
    ) +
    geom_point(data = rank_df, aes(year, rank, color = player), size = 3.2) +
    geom_text(
      data = ends,
      aes(year, rank, label = player, color = player),
      hjust = -0.35, size = 3.4, fontface = "bold", show.legend = FALSE
    ) +
    scale_color_manual(values = player_colors, guide = "none") +
    scale_y_reverse(breaks = 1:5) +
    scale_x_continuous(
      breaks = x_breaks,
      labels = x_labels,
      expand = expansion(mult = c(0.02, 0.09))
    ) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1)) +
    labs(title = title, x = NULL, y = "rank")
}

# x positions for the vault steps + final total on the rank charts
rank_steps <- c("postcards", "circus", "shares", "timetable", "total")
rank_step_x <- setNames(1901 + 3 * seq_along(rank_steps), rank_steps)
rank_breaks <- c(unique(scores$year), rank_step_x)
rank_labels <- c(unique(scores$year), "post cards", "circus flyer",
                 "shares", "timetable", "total")
```

```{r ranks-game}
#| fig-height: 4
ranks <- cum_scores |>
  group_by(year) |>
  mutate(
    game_rank = rank(-dollars, ties.method = "min"),
    campaign_rank = rank(-total, ties.method = "min")
  ) |>
  ungroup()

# per-component ranks: who earned the most from each vault addition,
# then the final standings as 'total'
comp_ranks <- dbGetQuery(con, "
  SELECT player, component, dollars FROM final_components
  WHERE component <> 'bankslips'
") |>
  group_by(component) |>
  mutate(rank = rank(-dollars, ties.method = "min")) |>
  ungroup() |>
  transmute(player, year = rank_step_x[component], rank)

total_rank <- dbGetQuery(con, "SELECT player, place FROM v_final_standings") |>
  transmute(player, year = rank_step_x[["total"]], rank = place)

game_ranks_full <- bind_rows(
  transmute(ranks, player = as.character(player), year, rank = game_rank),
  comp_ranks,
  total_rank
) |>
  mutate(player = factor(player, levels = names(player_colors)))

plot_bump(
  game_ranks_full,
  "Rank within each game, each vault addition, and the final total",
  rank_breaks, rank_labels
)
```

```{r ranks-campaign}
#| fig-height: 4
# cumulative rank at every year and after each vault addition steps in
cum_ranks_full <- cum_full |>
  bind_rows(
    cum_full |>
      filter(year == max(year)) |>
      mutate(year = rank_step_x[["total"]])
  ) |>
  group_by(year) |>
  mutate(rank = rank(-total, ties.method = "min")) |>
  ungroup() |>
  transmute(player, year, rank)

plot_bump(
  cum_ranks_full,
  "Rank on running total (bank slips, then each vault addition)",
  rank_breaks, rank_labels
)
```

## Where do the points come from?

Bank slips break each game's score into components: coins in hand,
bonus for trains remaining, and completed-ticket value.

```{r components}
components <- scores |>
  select(player, year, coins_in_hand, trains_bonus, ticket_value) |>
  pivot_longer(-c(player, year), names_to = "component", values_to = "value") |>
  mutate(component = recode(component,
    coins_in_hand = "coins in hand",
    trains_bonus = "trains remaining bonus",
    ticket_value = "ticket value"
  ))

component_totals <- components |>
  group_by(player, component) |>
  summarise(total = sum(value, na.rm = TRUE), .groups = "drop")

ggplot(component_totals, aes(player, total, fill = player)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = player_colors) +
  facet_wrap(~component) +
  labs(
    title = "Campaign totals by scoring component",
    x = NULL, y = "dollars"
  )
```

```{r components-time}
#| fig-height: 6
ggplot(components, aes(year, value, color = player)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = player_colors) +
  facet_wrap(~component, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Scoring components over time",
    x = NULL, y = "dollars", color = NULL
  )
```

The trains-remaining bonus can be read backwards: the bank-slip
schedule pays $16/12/9/7/6 for 0–4 trains left and $4/2/0 for 5–7,
8–10, 11+ — so the bonus reveals how many trains each company had
left on the table:

```{r trains-remaining}
#| fig-height: 4.5
trains <- dbGetQuery(con, "SELECT * FROM v_trains_remaining") |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    trains_plot = case_when(
      is.na(trains_min)          ~ NA_real_,
      is.na(trains_max)          ~ 11,
      TRUE                       ~ (trains_min + trains_max) / 2
    )
  )

ggplot(trains, aes(year, trains_plot, color = player)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_color_manual(values = player_colors) +
  scale_x_continuous(breaks = unique(scores$year)) +
  scale_y_continuous(
    breaks = c(0:4, 6, 9, 11),
    labels = c("0", "1", "2", "3", "4", "5-7", "8-10", "11+")
  ) +
  labs(
    title = "Trains remaining at game end (derived from the bonus)",
    x = NULL, y = "trains remaining", color = NULL
  )
```

One bank slip needed an explanation: yellow's 1883 bonus of $28 is off
the schedule — a postcard effect that game paid $28 instead of the
usual $16 for 0 trains, so it plots as 0 trains remaining.

## Tickets

Punches per player — how many ticket completions each company recorded
over the whole campaign, split by normal tickets and postcard tickets
(single-punch tickets that retire immediately and earn a post card).

```{r punches}
punches <- dbGetQuery(con, "
  SELECT p.color AS player, c.capacity, SUM(p.punches) AS punches
  FROM ticket_punches p
  JOIN tickets t ON t.ticket_id = p.ticket_id
  JOIN card_ref_capacity c ON c.ref_id = t.ref_id AND c.color = p.color
  GROUP BY p.color, c.capacity
") |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    type = ifelse(capacity == 1, "postcard tickets", "normal tickets")
  )

ggplot(punches, aes(player, punches, fill = player)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = punches),
    vjust = -0.4, size = 3, color = "gray30"
  ) +
  scale_fill_manual(values = player_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.09))) +
  facet_wrap(~type) +
  labs(
    title = "Ticket punches per company",
    x = NULL, y = "punches"
  )
```

Ticket dollars actually collected (ticket value × punches):

```{r ticket-dollars}
ticket_dollars <- dbGetQuery(con, "
  SELECT p.color AS player, SUM(p.punches * t.value) AS dollars
  FROM ticket_punches p
  JOIN tickets t ON t.ticket_id = p.ticket_id
  GROUP BY p.color
") |>
  mutate(player = factor(player, levels = names(player_colors)))

ggplot(ticket_dollars, aes(player, dollars, fill = player)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = player_colors) +
  labs(
    title = "Total dollars from completed tickets",
    x = NULL, y = "dollars"
  )
```

## Claims

The endgame scratch cards. On partially-scratched cards the scratched
spots are exactly the claims collected during play; fully-scratched cards
were opened out of curiosity afterwards, so their in-play earnings are
unknown.

```{r claims}
claims <- dbGetQuery(con, "SELECT * FROM v_claim_earnings")
knitr::kable(
  claims |>
    mutate(
      earned = ifelse(earnings_known == 1, earned_dollars, "unknown (fully scratched)"),
      owner = coalesce(owner, note, "?")
    ) |>
    select(card = card_code, town, owner, earned),
  caption = "Claim card earnings (all cards hide the same $98 total but in random ordering)"
)
```

## Company shares

Sixty shares across six railway companies (including the neutral white
railway), ten per company. Most shares in a company ranks first (at
least one share required to qualify; ties broken by the lowest-numbered
share held), paying $20/15/10/5 down the ranks.

```{r shares}
#| fig-height: 5
company_colors <- c(player_colors, white = "#9e9e9e", total = "#555555")

shares <- dbGetQuery(con, "SELECT * FROM shares")

share_totals <- shares |>
  group_by(player) |>
  summarise(qty = sum(qty), .groups = "drop") |>
  mutate(company = "total")

shares_plot <- bind_rows(shares, share_totals) |>
  group_by(company) |>
  mutate(alpha_val = ifelse(qty == 0, 0, 0.15 + 0.85 * qty / max(qty))) |>
  ungroup() |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    company = factor(company,
      levels = rev(c("black", "blue", "green", "yellow", "red", "white", "total"))
    )
  )

ggplot(shares_plot, aes(player, company)) +
  geom_tile(aes(fill = company, alpha = alpha_val),
    color = "white", linewidth = 1.2
  ) +
  geom_text(
    aes(label = ifelse(qty > 0, qty, ""),
        color = alpha_val > 0.6),
    size = 3.6, fontface = "bold"
  ) +
  scale_fill_manual(values = company_colors, guide = "none") +
  scale_alpha_identity() +
  scale_color_manual(values = c(`TRUE` = "white", `FALSE` = "gray25"),
    guide = "none"
  ) +
  coord_fixed(ratio = 0.6) +
  labs(
    title = "Share holdings",
    caption = "Each company has 10 shares.",
    x = "players", y = "company"
  )
```

```{r share-payouts}
#| fig-height: 3.5
payouts <- dbGetQuery(con, "SELECT * FROM share_payouts") |>
  mutate(player = factor(player, levels = names(player_colors)))

ggplot(payouts, aes(player, dollars, fill = player)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = dollars), vjust = -0.4, size = 3, color = "gray30") +
  scale_fill_manual(values = player_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  labs(
    title = "Share dollars ($295 paid out in total)",
    x = NULL, y = "dollars"
  )
```

Green's diversified portfolio (22 of 60 shares, holding stakes in every
company) pays $110 — more than a third of the entire pool.

## The board

The game board as a network: 200 tracks connecting 63 cities across 12
jigsaw sections. Fictional cities (Nuevos Angeles, Cemetery City,
Pacific Haven, Baja) sit at plausible approximate positions; dashed
gray = unbuilt track beds. The cities are colored by the sticker
locations of *our* campaign — your board may vary.

```{r board}
#| fig-height: 7
coords <- dbGetQuery(con, "
  SELECT cc.*, c.company, c.port
  FROM city_coords cc JOIN cities c ON c.city = cc.city
") |>
  mutate(
    port_type = ifelse(coalesce(port, 0) == 1, "port", "inland"),
    company = ifelse(is.na(company), NA, company)
  )
routes <- dbGetQuery(con, "
  SELECT r.*, a.lat AS lat_a, a.lon AS lon_a, b.lat AS lat_b, b.lon AS lon_b
  FROM routes r
  JOIN city_coords a ON a.city = r.city_a
  JOIN city_coords b ON b.city = r.city_b
")

track_palette <- c(
  black = "#1a1a1a", blue = "#1f77c4", green = "#2a9d3a",
  yellow = "#e6b400", red = "#d62728", white = "#c8c8c8",
  gray = "#8a8a8a"
)

sticker_colors <- c(
  player_colors,
  "Mama O'Connel" = "#8B5A2B",   # brown
  big = "#9467bd"                 # purple
)

ggplot() +
  geom_segment(
    data = filter(routes, is.na(track_color)),
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray70", linetype = "dashed", linewidth = 0.5
  ) +
  geom_segment(
    data = filter(routes, !is.na(track_color)),
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b,
        color = track_color, linewidth = length),
    alpha = 0.55
  ) +
  geom_point(
    data = coords,
    aes(lon, lat, fill = company, shape = port_type),
    size = 2.6, color = "gray20", stroke = 0.5
  ) +
  geom_text(
    data = coords, aes(lon, lat, label = city),
    size = 2.4, vjust = -1.0, check_overlap = TRUE
  ) +
  scale_color_manual(values = track_palette, guide = "none") +
  scale_fill_manual(
    values = sticker_colors, na.value = NA,
    name = "company sticker",
    guide = guide_legend(override.aes = list(shape = 21, size = 3))
  ) +
  scale_shape_manual(
    values = c(inland = 21, port = 24),
    name = NULL,
    guide = guide_legend(override.aes = list(size = 3))
  ) +
  scale_linewidth_continuous(range = c(0.3, 1.8), guide = "none") +
  coord_quickmap() +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(title = "The board: tracks as printed, cities by company sticker (triangles = ports)")
```

### Opening the frontiers

How the map grows (per the official rules): at the end of a game, the
winner chooses which jigsaw piece to add next — its cities and tickets
come into play from the *next* game on. The dead letter office
remembers the order: frontier cards sit between dated story markers in
the pile, giving both the sequence and (with the one-frontier-per-game
constraint) the year each section became playable. Drag the slider (or
press play) to watch the board unfold from the initial East Coast start
area; cities are colored by jigsaw section, and tracks appear once both
their endpoints are unlocked.

```{r frontier-unlock}
#| fig-height: 6.5
frontier_map <- c(
  "great plains" = "Great Plains", "florida" = "Florida",
  "sierra madre" = "Sierra Madre", "haunted waste" = "Haunted Wastes",
  "badlands" = "Badlands", "california" = "California",
  "open range" = "Open Range", "cascadia" = "Cascadia"
)

unlocks <- dbGetQuery(con, "
  SELECT s.id, s.name, w.latest_year
  FROM story_cards s
  JOIN v_retirement_window w ON w.run_id = s.id
") |>
  mutate(key = tolower(name)) |>
  filter(key %in% names(frontier_map)) |>
  arrange(desc(id)) |>   # deeper in the pile = chosen earlier
  transmute(section = unname(frontier_map[key]), year = latest_year) |>
  # a frontier is chosen at the end of one game and playable from the
  # next, so in-play years must strictly increase (one per game); this
  # also corrects two markers that sit scrambled in the pile
  mutate(year = Reduce(
    function(prev, cur) max(cur, prev + 3),
    year, accumulate = TRUE
  ))

initial_sections <- c("Northeast", "Atlantic Coast", "The South", "Appalachia")

city_sections <- dbGetQuery(con, "
  SELECT c.city, c.section, cc.lat, cc.lon FROM cities c
  JOIN city_coords cc ON cc.city = c.city
")

# cumulative frames: each step adds one frontier's sections
acc <- initial_sections
frame_defs <- list(tibble(frame = "1865 · Initial", section = acc))
for (i in seq_len(nrow(unlocks))) {
  acc <- c(acc, unlocks$section[i])
  frame_defs[[i + 1]] <- tibble(
    frame = paste0(unlocks$year[i], " · ", unlocks$section[i]),
    section = acc
  )
}
frames_df <- bind_rows(frame_defs)

cities_anim <- frames_df |>
  inner_join(city_sections, by = "section",
    relationship = "many-to-many"
  ) |>
  mutate(tooltip = paste0("<b>", city, "</b><br>", section))

routes_anim <- frames_df |>
  distinct(frame) |>
  cross_join(routes) |>
  semi_join(
    cities_anim |> select(frame, city),
    by = c("frame", "city_a" = "city")
  ) |>
  semi_join(
    cities_anim |> select(frame, city),
    by = c("frame", "city_b" = "city")
  )

p_unlock <- ggplot() +
  geom_segment(
    data = routes_anim,
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b, frame = frame),
    color = "gray80", linewidth = 0.4
  ) +
  geom_point(
    data = cities_anim,
    aes(lon, lat, color = section, frame = frame, text = tooltip),
    size = 2.4
  ) +
  coord_fixed(ratio = 1.25) +
  theme_void() +
  labs(title = "How the board unfolded", color = "jigsaw section")

ggplotly(p_unlock, tooltip = "text") |>
  animation_opts(frame = 1200, transition = 300, redraw = FALSE) |>
  animation_slider(currentvalue = list(prefix = "unlocked: ", font = list(size = 14))) |>
  config(displayModeBar = FALSE)
```

### Hotspots

Which cities did completed tickets actually connect? Each punch on a
ticket counts once for each of its two endpoint cities. **Hover a
city** for its name, punch count, and most frequent visitor — the
company that punched the most tickets through it.

```{r hotspots}
#| fig-height: 6.5
city_players <- dbGetQuery(con, "
  WITH endpoint_punches AS (
    SELECT t.from_city AS city, p.color AS player, p.punches FROM tickets t
    JOIN ticket_punches p ON p.ticket_id = t.ticket_id
    UNION ALL
    SELECT t.to_city, p.color, p.punches FROM tickets t
    JOIN ticket_punches p ON p.ticket_id = t.ticket_id
  )
  SELECT city, player, SUM(punches) AS punches
  FROM endpoint_punches GROUP BY city, player
")

top_visitors <- city_players |>
  filter(punches > 0) |>
  group_by(city) |>
  filter(punches == max(punches)) |>
  summarise(
    top_label = paste0(paste(player, collapse = " & "),
                       " (", first(punches), ")"),
    .groups = "drop"
  )

hot <- city_players |>
  group_by(city) |>
  summarise(punches = sum(punches), .groups = "drop") |>
  left_join(top_visitors, by = "city") |>
  inner_join(coords, by = "city") |>
  mutate(tooltip = paste0(
    "<b>", city, "</b><br>punches: ", punches,
    "<br>most frequent visitor: ", coalesce(top_label, "none")
  ))

p_hot <- ggplot() +
  geom_segment(
    data = routes,
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray85", linewidth = 0.4
  ) +
  geom_point(
    data = hot,
    aes(lon, lat, size = punches, color = punches, text = tooltip),
    alpha = 0.8
  ) +
  scale_size_continuous(range = c(1, 9)) +
  scale_color_viridis_c(option = "inferno", direction = -1) +
  coord_fixed(ratio = 1.25) +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "Ticket hotspots: endpoint cities weighted by punches")

ggplotly(p_hot, tooltip = "text") |>
  config(displayModeBar = FALSE)
```

### When was the network built?

Retired tickets sit between dated story markers in the dead letter
office, which bounds when they were completed. Using the midpoint of
each ticket's retirement window as a rough timestamp:

```{r eras}
#| fig-height: 5
era_tickets <- dbGetQuery(con, "
  SELECT t.from_city, t.to_city,
         (COALESCE(w.earliest_year, 1865) + COALESCE(w.latest_year, 1901)) / 2 AS mid_year,
         SUM(p.punches) AS punches
  FROM tickets t
  JOIN v_retirement_window w ON w.run_id = t.run_id
  JOIN ticket_punches p ON p.ticket_id = t.ticket_id
  WHERE t.active = 0
  GROUP BY t.ticket_id
  HAVING punches > 0
") |>
  mutate(era = cut(mid_year,
    breaks = c(1864, 1874, 1886, 1901),
    labels = c("early (1865-1874)", "mid (1877-1886)", "late (1889-1901)")
  )) |>
  inner_join(coords, by = c("from_city" = "city")) |>
  rename(lat_a = lat, lon_a = lon) |>
  inner_join(coords, by = c("to_city" = "city")) |>
  rename(lat_b = lat, lon_b = lon)

ggplot(era_tickets) +
  geom_segment(
    data = routes,
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray88", linewidth = 0.3
  ) +
  geom_segment(
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b,
        linewidth = punches),
    color = "#c1440e", alpha = 0.5
  ) +
  scale_linewidth_continuous(range = c(0.3, 1.6), guide = "none") +
  coord_quickmap() +
  facet_wrap(~era) +
  theme_void() +
  labs(
    title = "Completed ticket connections by campaign era",
    subtitle = "Straight lines connect ticket endpoints (not actual track paths)"
  )
```

## Next up

- Route the ticket connections along actual board tracks (shortest path
  over the `routes` network) instead of straight lines.
- Claim / postcard geography: where the postcard tickets cluster.

```{r close}
#| include: false
dbDisconnect(con)
```
