---
title: "Ticket to Ride Legacy — Data Dive"
subtitle: "One campaign, 12 plays, 5 companies — reconstructed from the dead letter office"
---

```{r setup}
#| include: false
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)

con <- dbConnect(SQLite(), "ttr.db")

player_colors <- c(
  black  = "#1a1a1a",
  blue   = "#1f77c4",
  green  = "#2a9d3a",
  yellow = "#e6b400",
  red    = "#d62728"
)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold")
    )
)

scores <- dbGetQuery(con, "SELECT * FROM v_scores") |>
  mutate(player = factor(player, levels = names(player_colors)))

players <- dbGetQuery(con, "SELECT * FROM players")
```

## Score progression

Dollars earned per game, per company. Each game is one year of the
campaign, 1865–1898. The dashed gray line is the suggested dollars for
a player missing in action that game — a baseline value.

```{r per-game}
missing_scores <- dbGetQuery(con, "
  SELECT year, dollars FROM bank_slips WHERE player = 'missing'
") |>
  mutate(player = "(missing)")

pg_levels <- c(names(player_colors), "(missing)")
pg_colors <- c(player_colors, "(missing)" = "gray45")
pg_lines <- setNames(c(rep("solid", 5), "dashed"), pg_levels)

per_game <- bind_rows(
  mutate(scores, player = as.character(player)),
  missing_scores
) |>
  mutate(player = factor(player, levels = pg_levels))

ggplot(per_game, aes(year, dollars, color = player, linetype = player)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(per_game, player != "(missing)"), size = 2) +
  scale_color_manual(values = pg_colors) +
  scale_linetype_manual(values = pg_lines) +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Dollars per game",
    x = NULL, y = "dollars", color = NULL, linetype = NULL
  )
```

The campaign total is what decides the winner — the running sum of the
bank slips through 1898, then each post-campaign addition stepped in
one at a time: post cards (from the vaults), circus flyer, shares, and
finally the timetable. The number at the right is each company's final
score.

```{r cumulative}
#| fig-height: 6
cum_scores <- scores |>
  arrange(player, year) |>
  group_by(player) |>
  mutate(total = cumsum(dollars)) |>
  ungroup()

component_order <- c("postcards", "circus", "shares", "timetable")
component_x <- setNames(1901 + 3 * seq_along(component_order), component_order)

additions_wide <- dbGetQuery(con, "
  SELECT player, component, dollars FROM final_components
  WHERE component <> 'bankslips'
")

steps <- cum_scores |>
  filter(year == max(year)) |>
  select(player, total)
step_rows <- list()
for (comp in component_order) {
  steps <- steps |>
    left_join(filter(additions_wide, component == comp) |>
        select(player, add = dollars), by = "player") |>
    mutate(total = total + coalesce(add, 0)) |>
    select(-add)
  step_rows[[comp]] <- mutate(steps, year = component_x[[comp]])
}

cum_full <- bind_rows(
  select(cum_scores, player, year, total),
  bind_rows(step_rows)
) |>
  mutate(player = factor(player, levels = names(player_colors)))

totals_lab <- cum_full |>
  filter(year == max(year)) |>
  arrange(total) |>
  mutate(
    lab_y = total,
    # push overlapping labels apart (min 45 dollar-units of separation)
    lab_y = Reduce(
      function(prev, cur) max(cur, prev + 45),
      lab_y, accumulate = TRUE
    )
  )

ggplot(cum_full, aes(year, total, color = player)) +
  annotate("segment",
    x = 1865, xend = 1898, y = -95, yend = -95,
    color = "gray55", linewidth = 0.5
  ) +
  annotate("text",
    x = (1865 + 1898) / 2, y = -160, label = "bank slips",
    color = "gray45", size = 3.2
  ) +
  annotate("segment",
    x = min(component_x), xend = max(component_x), y = -95, yend = -95,
    color = "gray55", linewidth = 0.5
  ) +
  annotate("text",
    x = mean(range(component_x)), y = -160, label = "vault",
    color = "gray45", size = 3.2
  ) +
  geom_vline(xintercept = 1901, color = "gray80", linetype = "dotted") +
  annotate("text",
    x = 1900, y = 650, label = "post campaign",
    angle = 90, color = "gray55", size = 3
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_text(
    data = totals_lab,
    aes(y = lab_y, label = paste0("$", total)),
    hjust = -0.2, size = 3.3, fontface = "bold", show.legend = FALSE
  ) +
  scale_color_manual(values = player_colors) +
  scale_x_continuous(
    breaks = c(unique(scores$year), component_x),
    labels = c(unique(scores$year), "post cards", "circus flyer",
               "shares", "timetable"),
    expand = expansion(mult = c(0.02, 0.06))
  ) +
  coord_cartesian(clip = "off") +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    plot.margin = margin(10, 20, 26, 10)
  ) +
  labs(
    title = "Campaign total (bank slips, then each addition stepped in)",
    x = NULL, y = "cumulative dollars", color = NULL
  )
```

The post-campaign vault additions nearly rewrote the podium: green's $259 of
additions ($110 shares, $71 circus, $30 timetable, $48 post cards)
closed the gap to black from $66 to just **$18** — final score $1363 to
$1345.

```{r final-additions}
#| fig-height: 4
additions <- dbGetQuery(con, "
  SELECT player, component, dollars
  FROM final_components
  WHERE component <> 'bankslips'
") |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    component = factor(component,
      levels = c("shares", "circus", "timetable", "postcards")
    )
  )

ggplot(additions, aes(player, dollars, fill = component)) +
  geom_col(color = "white", linewidth = 0.3) +
  scale_fill_manual(values = c(
    shares = "#4477aa", circus = "#ee6677",
    timetable = "#228833", postcards = "#ccbb44"
  )) +
  labs(
    title = "Post-campaign vault additions by component",
    x = NULL, y = "dollars", fill = NULL
  )
```

### Did the ordering change?

Rank 1 = leading. First the rank within each individual game — extended
past 1898 with the rank within each vault addition (who earned the most
post cards, circus flyer, shares, timetable) and the final total. Then
the rank on the running campaign total as each of those steps in — the
race that matters.

```{r rank-helpers}
#| include: false
# sigmoid interpolation between rank points, sankey/bump style
bump_curves <- function(df, n = 40) {
  df <- arrange(df, player, year)
  do.call(rbind, lapply(split(df, df$player), function(d) {
    segs <- lapply(seq_len(nrow(d) - 1), function(i) {
      t <- seq(0, 1, length.out = n)
      s <- 1 / (1 + exp(-12 * (t - 0.5)))   # sigmoid easing
      data.frame(
        player = d$player[i],
        x = d$year[i] + t * (d$year[i + 1] - d$year[i]),
        y = d$rank[i] + s * (d$rank[i + 1] - d$rank[i])
      )
    })
    do.call(rbind, segs)
  }))
}

plot_bump <- function(rank_df, title, x_breaks, x_labels) {
  curves <- bump_curves(rank_df)
  ends <- filter(rank_df, year == max(year))
  ggplot() +
    geom_vline(xintercept = 1901, color = "gray80", linetype = "dotted") +
    geom_line(
      data = curves, aes(x, y, color = player, group = player),
      linewidth = 2.4, alpha = 0.85, lineend = "round"
    ) +
    geom_point(data = rank_df, aes(year, rank, color = player), size = 3.2) +
    geom_text(
      data = ends,
      aes(year, rank, label = player, color = player),
      hjust = -0.35, size = 3.4, fontface = "bold", show.legend = FALSE
    ) +
    scale_color_manual(values = player_colors, guide = "none") +
    scale_y_reverse(breaks = 1:5) +
    scale_x_continuous(
      breaks = x_breaks,
      labels = x_labels,
      expand = expansion(mult = c(0.02, 0.09))
    ) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1)) +
    labs(title = title, x = NULL, y = "rank")
}

# x positions for the vault steps + final total on the rank charts
rank_steps <- c("postcards", "circus", "shares", "timetable", "total")
rank_step_x <- setNames(1901 + 3 * seq_along(rank_steps), rank_steps)
rank_breaks <- c(unique(scores$year), rank_step_x)
rank_labels <- c(unique(scores$year), "post cards", "circus flyer",
                 "shares", "timetable", "total")
```

```{r ranks-game}
#| fig-height: 4
ranks <- cum_scores |>
  group_by(year) |>
  mutate(
    game_rank = rank(-dollars, ties.method = "min"),
    campaign_rank = rank(-total, ties.method = "min")
  ) |>
  ungroup()

# per-component ranks: who earned the most from each vault addition,
# then the final standings as 'total'
comp_ranks <- dbGetQuery(con, "
  SELECT player, component, dollars FROM final_components
  WHERE component <> 'bankslips'
") |>
  group_by(component) |>
  mutate(rank = rank(-dollars, ties.method = "min")) |>
  ungroup() |>
  transmute(player, year = rank_step_x[component], rank)

total_rank <- dbGetQuery(con, "SELECT player, place FROM v_final_standings") |>
  transmute(player, year = rank_step_x[["total"]], rank = place)

game_ranks_full <- bind_rows(
  transmute(ranks, player = as.character(player), year, rank = game_rank),
  comp_ranks,
  total_rank
) |>
  mutate(player = factor(player, levels = names(player_colors)))

plot_bump(
  game_ranks_full,
  "Rank within each game, each vault addition, and the final total",
  rank_breaks, rank_labels
)
```

```{r ranks-campaign}
#| fig-height: 4
# cumulative rank at every year and after each vault addition steps in
cum_ranks_full <- cum_full |>
  bind_rows(
    cum_full |>
      filter(year == max(year)) |>
      mutate(year = rank_step_x[["total"]])
  ) |>
  group_by(year) |>
  mutate(rank = rank(-total, ties.method = "min")) |>
  ungroup() |>
  transmute(player, year, rank)

plot_bump(
  cum_ranks_full,
  "Rank on running total (bank slips, then each vault addition)",
  rank_breaks, rank_labels
)
```

## Where do the points come from?

Bank slips break each game's score into components: coins in hand,
bonus for trains remaining, and completed-ticket value.

```{r components}
components <- scores |>
  select(player, year, coins_in_hand, trains_bonus, ticket_value) |>
  pivot_longer(-c(player, year), names_to = "component", values_to = "value") |>
  mutate(component = recode(component,
    coins_in_hand = "coins in hand",
    trains_bonus = "trains remaining bonus",
    ticket_value = "ticket value"
  ))

component_totals <- components |>
  group_by(player, component) |>
  summarise(total = sum(value, na.rm = TRUE), .groups = "drop")

ggplot(component_totals, aes(player, total, fill = player)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = player_colors) +
  facet_wrap(~component) +
  labs(
    title = "Campaign totals by scoring component",
    x = NULL, y = "dollars"
  )
```

```{r components-time}
#| fig-height: 6
ggplot(components, aes(year, value, color = player)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = player_colors) +
  facet_wrap(~component, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Scoring components over time",
    x = NULL, y = "dollars", color = NULL
  )
```

The trains-remaining bonus can be read backwards: the bank-slip
schedule pays $16/12/9/7/6 for 0–4 trains left and $4/2/0 for 5–7,
8–10, 11+ — so the bonus reveals how many trains each company had
left on the table:

```{r trains-remaining}
#| fig-height: 4.5
trains <- dbGetQuery(con, "SELECT * FROM v_trains_remaining") |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    trains_plot = case_when(
      is.na(trains_min)          ~ NA_real_,
      is.na(trains_max)          ~ 11,
      TRUE                       ~ (trains_min + trains_max) / 2
    )
  )

ggplot(trains, aes(year, trains_plot, color = player)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_color_manual(values = player_colors) +
  scale_x_continuous(breaks = unique(scores$year)) +
  scale_y_continuous(
    breaks = c(0:4, 6, 9, 11),
    labels = c("0", "1", "2", "3", "4", "5-7", "8-10", "11+")
  ) +
  labs(
    title = "Trains remaining at game end (derived from the bonus)",
    x = NULL, y = "trains remaining", color = NULL
  )
```

One bank slip needed an explanation: yellow's 1883 bonus of $28 is off
the schedule — a postcard effect that game paid $28 instead of the
usual $16 for 0 trains, so it plots as 0 trains remaining.

## Tickets

Punches per player — how many ticket completions each company recorded
over the whole campaign, split by normal tickets and postcard tickets
(single-punch tickets that retire immediately and earn a post card).

```{r punches}
punches <- dbGetQuery(con, "
  SELECT p.color AS player, c.capacity, SUM(p.punches) AS punches
  FROM ticket_punches p
  JOIN tickets t ON t.ticket_id = p.ticket_id
  JOIN card_ref_capacity c ON c.ref_id = t.ref_id AND c.color = p.color
  GROUP BY p.color, c.capacity
") |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    type = ifelse(capacity == 1, "postcard tickets", "normal tickets")
  )

ggplot(punches, aes(player, punches, fill = player)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = punches),
    vjust = -0.4, size = 3, color = "gray30"
  ) +
  scale_fill_manual(values = player_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.09))) +
  facet_wrap(~type) +
  labs(
    title = "Ticket punches per company",
    x = NULL, y = "punches"
  )
```

Ticket dollars actually collected (ticket value × punches):

```{r ticket-dollars}
ticket_dollars <- dbGetQuery(con, "
  SELECT p.color AS player, SUM(p.punches * t.value) AS dollars
  FROM ticket_punches p
  JOIN tickets t ON t.ticket_id = p.ticket_id
  GROUP BY p.color
") |>
  mutate(player = factor(player, levels = names(player_colors)))

ggplot(ticket_dollars, aes(player, dollars, fill = player)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = player_colors) +
  labs(
    title = "Total dollars from completed tickets",
    x = NULL, y = "dollars"
  )
```

## Claims

The endgame scratch cards. On partially-scratched cards the scratched
spots are exactly the claims collected during play; fully-scratched cards
were opened out of curiosity afterwards, so their in-play earnings are
unknown.

```{r claims}
claims <- dbGetQuery(con, "SELECT * FROM v_claim_earnings")
knitr::kable(
  claims |>
    mutate(
      earned = ifelse(earnings_known == 1, earned_dollars, "unknown (fully scratched)"),
      owner = coalesce(owner, note, "?")
    ) |>
    select(card = card_code, town, owner, earned),
  caption = "Claim card earnings (all cards hide the same $98 total but in random ordering)"
)
```

## Company shares

Sixty shares across six railway companies (including the neutral white
railway), ten per company. Most shares in a company ranks first (at
least one share required to qualify; ties broken by the lowest-numbered
share held), paying $20/15/10/5 down the ranks.

```{r shares}
#| fig-height: 5
company_colors <- c(player_colors, white = "#9e9e9e", total = "#555555")

shares <- dbGetQuery(con, "SELECT * FROM shares")

share_totals <- shares |>
  group_by(player) |>
  summarise(qty = sum(qty), .groups = "drop") |>
  mutate(company = "total")

shares_plot <- bind_rows(shares, share_totals) |>
  group_by(company) |>
  mutate(alpha_val = ifelse(qty == 0, 0, 0.15 + 0.85 * qty / max(qty))) |>
  ungroup() |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    company = factor(company,
      levels = rev(c("black", "blue", "green", "yellow", "red", "white", "total"))
    )
  )

ggplot(shares_plot, aes(player, company)) +
  geom_tile(aes(fill = company, alpha = alpha_val),
    color = "white", linewidth = 1.2
  ) +
  geom_text(
    aes(label = ifelse(qty > 0, qty, ""),
        color = alpha_val > 0.6),
    size = 3.6, fontface = "bold"
  ) +
  scale_fill_manual(values = company_colors, guide = "none") +
  scale_alpha_identity() +
  scale_color_manual(values = c(`TRUE` = "white", `FALSE` = "gray25"),
    guide = "none"
  ) +
  coord_fixed(ratio = 0.6) +
  labs(
    title = "Share holdings",
    caption = "Each company has 10 shares.",
    x = "players", y = "company"
  )
```

```{r share-payouts}
#| fig-height: 3.5
payouts <- dbGetQuery(con, "SELECT * FROM share_payouts") |>
  mutate(player = factor(player, levels = names(player_colors)))

ggplot(payouts, aes(player, dollars, fill = player)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = dollars), vjust = -0.4, size = 3, color = "gray30") +
  scale_fill_manual(values = player_colors) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  labs(
    title = "Share dollars ($295 paid out in total)",
    x = NULL, y = "dollars"
  )
```

Green's diversified portfolio (22 of 60 shares, holding stakes in every
company) pays $110 — more than a third of the entire pool.

## The board

The game board as a network: 200 tracks connecting 63 cities across 12
jigsaw sections. Fictional cities (Nuevos Angeles, Cemetery City,
Pacific Haven, Baja) sit at plausible approximate positions; dashed
gray = unbuilt track beds.

```{r board}
#| fig-height: 7
coords <- dbGetQuery(con, "
  SELECT cc.*, c.company, c.port
  FROM city_coords cc JOIN cities c ON c.city = cc.city
") |>
  mutate(
    port_type = ifelse(coalesce(port, 0) == 1, "port", "inland"),
    company = ifelse(is.na(company), NA, company)
  )
routes <- dbGetQuery(con, "
  SELECT r.*, a.lat AS lat_a, a.lon AS lon_a, b.lat AS lat_b, b.lon AS lon_b
  FROM routes r
  JOIN city_coords a ON a.city = r.city_a
  JOIN city_coords b ON b.city = r.city_b
")

track_palette <- c(
  black = "#1a1a1a", blue = "#1f77c4", green = "#2a9d3a",
  yellow = "#e6b400", red = "#d62728", white = "#c8c8c8",
  gray = "#8a8a8a"
)

sticker_colors <- c(
  player_colors,
  "Mama O'Connel" = "#8B5A2B",   # brown
  big = "#9467bd"                 # purple
)

ggplot() +
  geom_segment(
    data = filter(routes, is.na(track_color)),
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray70", linetype = "dashed", linewidth = 0.5
  ) +
  geom_segment(
    data = filter(routes, !is.na(track_color)),
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b,
        color = track_color, linewidth = length),
    alpha = 0.55
  ) +
  geom_point(
    data = coords,
    aes(lon, lat, fill = company, shape = port_type),
    size = 2.6, color = "gray20", stroke = 0.5
  ) +
  geom_text(
    data = coords, aes(lon, lat, label = city),
    size = 2.4, vjust = -1.0, check_overlap = TRUE
  ) +
  scale_color_manual(values = track_palette, guide = "none") +
  scale_fill_manual(
    values = sticker_colors, na.value = NA,
    name = "company sticker",
    guide = guide_legend(override.aes = list(shape = 21, size = 3))
  ) +
  scale_shape_manual(
    values = c(inland = 21, port = 24),
    name = NULL,
    guide = guide_legend(override.aes = list(size = 3))
  ) +
  scale_linewidth_continuous(range = c(0.3, 1.8), guide = "none") +
  coord_quickmap() +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(title = "The board: tracks as printed, cities by company sticker (triangles = ports)")
```

### Hotspots

Which cities did completed tickets actually connect? Each punch on a
ticket counts once for each of its two endpoint cities.

```{r hotspots}
#| fig-height: 7
hot <- dbGetQuery(con, "
  WITH endpoint_punches AS (
    SELECT t.from_city AS city, p.punches FROM tickets t
    JOIN ticket_punches p ON p.ticket_id = t.ticket_id
    UNION ALL
    SELECT t.to_city, p.punches FROM tickets t
    JOIN ticket_punches p ON p.ticket_id = t.ticket_id
  )
  SELECT city, SUM(punches) AS punches
  FROM endpoint_punches GROUP BY city
") |>
  inner_join(coords, by = "city")

ggplot() +
  geom_segment(
    data = routes,
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray85", linewidth = 0.4
  ) +
  geom_point(
    data = hot,
    aes(lon, lat, size = punches, color = punches),
    alpha = 0.8
  ) +
  geom_text(
    data = slice_max(hot, punches, n = 12),
    aes(lon, lat, label = paste0(city, " (", punches, ")")),
    size = 2.8, vjust = -1.1, fontface = "bold", check_overlap = TRUE
  ) +
  scale_size_continuous(range = c(1, 10)) +
  scale_color_viridis_c(option = "inferno", direction = -1) +
  coord_quickmap() +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "Ticket hotspots: endpoint cities weighted by punches")
```

### When was the network built?

Retired tickets sit between dated story markers in the dead letter
office, which bounds when they were completed. Using the midpoint of
each ticket's retirement window as a rough timestamp:

```{r eras}
#| fig-height: 5
era_tickets <- dbGetQuery(con, "
  SELECT t.from_city, t.to_city,
         (COALESCE(w.earliest_year, 1865) + COALESCE(w.latest_year, 1901)) / 2 AS mid_year,
         SUM(p.punches) AS punches
  FROM tickets t
  JOIN v_retirement_window w ON w.run_id = t.run_id
  JOIN ticket_punches p ON p.ticket_id = t.ticket_id
  WHERE t.active = 0
  GROUP BY t.ticket_id
  HAVING punches > 0
") |>
  mutate(era = cut(mid_year,
    breaks = c(1864, 1874, 1886, 1901),
    labels = c("early (1865-1874)", "mid (1877-1886)", "late (1889-1901)")
  )) |>
  inner_join(coords, by = c("from_city" = "city")) |>
  rename(lat_a = lat, lon_a = lon) |>
  inner_join(coords, by = c("to_city" = "city")) |>
  rename(lat_b = lat, lon_b = lon)

ggplot(era_tickets) +
  geom_segment(
    data = routes,
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b),
    color = "gray88", linewidth = 0.3
  ) +
  geom_segment(
    aes(x = lon_a, y = lat_a, xend = lon_b, yend = lat_b,
        linewidth = punches),
    color = "#c1440e", alpha = 0.5
  ) +
  scale_linewidth_continuous(range = c(0.3, 1.6), guide = "none") +
  coord_quickmap() +
  facet_wrap(~era) +
  theme_void() +
  labs(
    title = "Completed ticket connections by campaign era",
    subtitle = "Straight lines connect ticket endpoints (not actual track paths)"
  )
```

## Next up

- Route the ticket connections along actual board tracks (shortest path
  over the `routes` network) instead of straight lines.
- Claim / postcard geography: where the postcard tickets cluster.

```{r close}
#| include: false
dbDisconnect(con)
```
