---
title: "Ticket to Ride Legacy — Data Dive"
subtitle: "One campaign, 13 games, 5 companies — reconstructed from the dead letter office"
---

```{r setup}
#| include: false
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)

con <- dbConnect(SQLite(), "ttr.db")

player_colors <- c(
  black  = "#1a1a1a",
  blue   = "#1f77c4",
  green  = "#2a9d3a",
  yellow = "#e6b400",
  red    = "#d62728"
)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold")
    )
)

scores <- dbGetQuery(con, "SELECT * FROM v_scores") |>
  mutate(player = factor(player, levels = names(player_colors)))

players <- dbGetQuery(con, "SELECT * FROM players")
```

## Score progression

Dollars earned per game, per company. Each game is one year of the
campaign, 1865–1898.

```{r per-game}
ggplot(scores, aes(year, dollars, color = player)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = player_colors) +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Dollars per game",
    x = NULL, y = "dollars", color = NULL
  )
```

The campaign total is what decides the winner — the running sum of all
game scores:

```{r cumulative}
cum_scores <- scores |>
  arrange(player, year) |>
  group_by(player) |>
  mutate(total = cumsum(dollars)) |>
  ungroup()

ggplot(cum_scores, aes(year, total, color = player)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = player_colors) +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Campaign total (cumulative dollars)",
    x = NULL, y = "cumulative dollars", color = NULL
  )
```

### Did the ordering change?

Rank 1 = leading. Left: rank within each game. Right: rank on
cumulative campaign total — the race that matters.

```{r ranks}
#| fig-height: 4.5
ranks <- cum_scores |>
  group_by(year) |>
  mutate(
    game_rank = rank(-dollars, ties.method = "min"),
    campaign_rank = rank(-total, ties.method = "min")
  ) |>
  ungroup() |>
  pivot_longer(c(game_rank, campaign_rank),
    names_to = "kind", values_to = "rank"
  ) |>
  mutate(kind = recode(kind,
    game_rank = "per game",
    campaign_rank = "campaign total"
  ))

ggplot(ranks, aes(year, rank, color = player)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = player_colors) +
  scale_y_reverse(breaks = 1:5) +
  scale_x_continuous(breaks = unique(scores$year)) +
  facet_wrap(~kind) +
  labs(title = "Ranking over the campaign", x = NULL, y = "rank", color = NULL)
```

## Where do the points come from?

Bank slips break each game's score into components: coins in hand,
bonus for trains remaining, and completed-ticket value.

```{r components}
components <- scores |>
  select(player, year, coins_in_hand, trains_bonus, ticket_value) |>
  pivot_longer(-c(player, year), names_to = "component", values_to = "value") |>
  mutate(component = recode(component,
    coins_in_hand = "coins in hand",
    trains_bonus = "trains remaining bonus",
    ticket_value = "ticket value"
  ))

component_totals <- components |>
  group_by(player, component) |>
  summarise(total = sum(value, na.rm = TRUE), .groups = "drop")

ggplot(component_totals, aes(player, total, fill = player)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = player_colors) +
  facet_wrap(~component, scales = "free_y") +
  labs(
    title = "Campaign totals by scoring component",
    x = NULL, y = "dollars"
  )
```

```{r components-time}
#| fig-height: 6
ggplot(components, aes(year, value, color = player)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = player_colors) +
  facet_wrap(~component, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Scoring components over time",
    x = NULL, y = "dollars", color = NULL
  )
```

## Tickets

Punches per player — how many ticket completions each company recorded
over the whole campaign, split by normal tickets and postcard tickets
(single-punch tickets that retire immediately and earn a post card).

```{r punches}
punches <- dbGetQuery(con, "
  SELECT p.color AS player, c.capacity, SUM(p.punches) AS punches
  FROM ticket_punches p
  JOIN tickets t ON t.ticket_id = p.ticket_id
  JOIN card_ref_capacity c ON c.ref_id = t.ref_id AND c.color = p.color
  GROUP BY p.color, c.capacity
") |>
  mutate(
    player = factor(player, levels = names(player_colors)),
    type = ifelse(capacity == 1, "postcard tickets", "normal tickets")
  )

ggplot(punches, aes(player, punches, fill = player)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = player_colors) +
  facet_wrap(~type) +
  labs(
    title = "Ticket punches per company",
    x = NULL, y = "punches"
  )
```

Ticket dollars actually collected (ticket value × punches):

```{r ticket-dollars}
ticket_dollars <- dbGetQuery(con, "
  SELECT p.color AS player, SUM(p.punches * t.value) AS dollars
  FROM ticket_punches p
  JOIN tickets t ON t.ticket_id = p.ticket_id
  GROUP BY p.color
") |>
  mutate(player = factor(player, levels = names(player_colors)))

ggplot(ticket_dollars, aes(player, dollars, fill = player)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = player_colors) +
  labs(
    title = "Total dollars from completed tickets",
    x = NULL, y = "dollars"
  )
```

## Claims

The endgame scratch cards. On partially-scratched cards the scratched
spots are exactly the claims collected during play; fully-scratched cards
were opened out of curiosity afterwards, so their in-play earnings are
unknown.

```{r claims}
claims <- dbGetQuery(con, "SELECT * FROM v_claim_earnings")
knitr::kable(
  claims |>
    mutate(
      earned = ifelse(earnings_known == 1, earned_dollars, "unknown (fully scratched)"),
      owner = ifelse(is.na(owner), "?", owner)
    ) |>
    select(card = card_code, town, owner, earned),
  caption = "Claim card earnings (all cards hide the same 98$ total)"
)
```

## Gender

Black and blue are played by men, green, yellow and red by women (no
names tracked). Average dollars per game by gender:

```{r gender}
#| fig-height: 4
gender_scores <- scores |>
  left_join(players, by = c("player" = "color"))

gender_summary <- gender_scores |>
  group_by(gender, year) |>
  summarise(mean_dollars = mean(dollars), .groups = "drop")

ggplot(gender_summary, aes(year, mean_dollars, color = gender)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c(male = "#4477aa", female = "#ee6677")) +
  scale_x_continuous(breaks = unique(scores$year)) +
  labs(
    title = "Average dollars per game by gender",
    x = NULL, y = "mean dollars", color = NULL
  )
```

```{r gender-test}
t_res <- t.test(dollars ~ gender, data = gender_scores)
```

A Welch t-test on per-game scores gives p = `r round(t_res$p.value, 3)`
(mean female `r round(t_res$estimate[1], 1)`, mean male
`r round(t_res$estimate[2], 1)`) — `r if (t_res$p.value < 0.05) "a statistically significant difference" else "no statistically significant difference"`
at the 5% level, though with 5 players and 12 games this is descriptive at
best.

## Next up

- Map the game board: city coordinates, route overlays, and completion
  "hotspots" weighted by punches.
- Use the dead-letter-office retirement windows (`v_ticket_retirement`)
  to animate *when* tracks were laid across the map.

```{r close}
#| include: false
dbDisconnect(con)
```
